Neuroelectrophysiology-Compatible Electrolytic Lesioning

Version 1

Abstract

Abstract

Lesion studies have historically been instrumental for establishing causal connections between brain and behavior. They stand to provide additional insight if integrated with multielectrode techniques common in systems neuroscience. Here we present and test a platform for creating electrolytic lesions through chronically implanted, intracortical multielectrode probes without compromising the ability to acquire neuroelectrophysiology. A custom-built current source provides stable current and allows for controlled, repeatable lesions in awake-behaving animals. Performance of this novel lesioning technique was validated using histology from ex vivo and in vivo testing, current and voltage traces from the device, and measurements of spiking activity before and after lesioning. This electrolytic lesioning method avoids disruptive procedures, provides millimeter precision over the extent and submilimeter precision over the location of the injury, and permits electrophysiological recording of single-unit activity from the remaining neuronal population after lesioning. This technique can be used in many areas of cortex, in several species, and theoretically with any multielectrode probe. The low-cost, external lesioning device can also easily be adopted into an existing electrophysiology recording setup. This technique is expected to enable future causal investigations of the recorded neuronal population’s role in neuronal circuit function, while simultaneously providing new insight into local reorganization after neuron loss.

Introduction

Neuroelectrophysiology–using electrodes to record the electrical signals generated by neurons–has been the defining technique that shifted neuroscience from macroscale anatomy and loss-of-function studies down to the microscale activity of individual neurons and synapses in a given region of interest Sporns (2016). Prior to the last few decades, electrophysiology was limited to simultaneous recordings of at most five neurons with small numbers of electrodes (typically one, two, or four) McNaughton et al. (1983). The development of neuroelectrophysiology recording techniques with a large number of electrodes starting in the 1970s Wise et al. (1970); Campbell et al. (1990, 1991); Rios et al. (2016); Hong and Lieber (2019) was a boon to systems neuroscience, enabling simultaneous recording from hundreds of neurons and revealing previously unseen aspects of population coding Cunningham and Yu (2014); Kalaska (2019); Shenoy and Kao (2021). Over the last five years, electrode count has continued to increase, with new probes containing thousands of electrodes Jun et al. (2017); Steinmetz et al. (2021); Obaid et al. (2020); Musk (2019). While experiments and analysis have revealed population activity that correlates strongly with behavioral output Yu et al. (2009); Churchland et al. (2012); Sadtler et al. (2014); Golub et al. (2018); Elsayed et al. (2016); Goldman et al. (2019); Gallego et al. (2020); Liu et al. (2020); Rasmussen et al. (2017), novel tools that both record from and inactivate neurons are required to establish causal connections to behavior Vaidya et al. (2019); Shenoy and Kao (2021); Slonina et al. (2022). Although many inactivation methods exist, to date it has been challenging to find a repeatable, long-term inactivation technique compatible with chronic intracortical neuroelectrophysiology.

Design Considerations and Existing Inactivation Methods

Three main design considerations are required for a technique that successfully combines electrophysiology with inactivation (further detailed in Appendix 1).

Stable electrophysiology pre- and post-inactivation: Avoiding physical disruption enables direct comparison of pre-lesion activity with both the acute and chronic stages of injury.

Ability to localize and control the size of the inactivation: Precise focal inactivations strike a balance between being large enough to alter performance but small enough to spare sufficient tissue to record local re-organization and recovery.

Cross-compatibility: A technique that can be used in many areas of cortex, with any multielectrode probe, and in several species will enable causal investigation across a large variety of contexts. A method that works across-species (especially in large-animals like rhesus macaques) would leverage the existing injury literature, recording technologies, and behavioral assays from rodents to new world monkeys, while also being well-suited for the behavioral sophistication and human homology of macaques Higo (2021).

Neuronal activity can either be temporarily or permanently inactivated, defined as manipulation or termination, respectively Vaidya et al. (2019). For clarity, termination refers to a technique that causes death of neurons, removing them from the circuit. We non-exhaustively review existing manipulation and termination techniques in Appendix 2. However, none of these existing techniques meet all of the above design considerations.

Manipulations enable a causal understanding of the relationship between neuronal activity and behavior by studying adaptation to and from a perturbation Slonina et al. (2022). Existing temporary inactivation methods include intracortical microstimulation, optogenetics, pharmacology, transcranial stimulation, cooling loops, and chemogenetics. Since neurons remain anatomically and physiologically viable, manipulations can be easily repeated, until desensitization occurs.

As termination causes cell death, it can generate stronger causal evidence than a manipulation Vaidya et al. (2019). For example, even though a transient manipulation of neuronal activity may temporarily disrupt behavior, only a sustained manipulation could elicit the system’s long-term adaption Slonina et al. (2022). Over the days Ferrier and Yeo (1884) and potentially weeks Bundy and Nudo (2019); Zeiler et al. (2015) following a termination, the surrounding circuitry and broader network may adapt, leading to behavioral recovery and demonstrating that although the terminated region was causally implicated in behavioral control in the moment, the terminated neurons are not themselves causally necessary. A sustained manipulation could accomplish a similar effect to a termination at the systems level, up to the point at which the manipulation stopped. However, it can be difficult to create sustained inactivation with existing reversible manipulation methods, limiting their ability to study the brain’s natural reorganization over timescales of days to months.

Termination methods overcome the problem of sustained and consistent inactivation from which temporary inactivation techniques suffer. Therefore, they enable a form of causal inference not possible with temporary inactivation methods Vaidya et al. (2019); Shenoy and Kao (2021). Existing termination techniques include mechanical trauma, endovascular occlusion, Rose Bengal mediated photothrombosis, and chemical lesioning.

Electrolytic Lesioning Through A Microelectrode Array

In order to best meet the three design considerations, we created a device to make an electrolytic lesion through the same microelectrode array used to record neuronal activity. The dual use of this microelectrode array achieves the first design constraint of stable electrophysiology pre- and post-inactivation, as it enables recording from the exact same signal source with minimal physical disruption or displacement. Aside from the initial implantation of the array, there are no invasive procedures required, removing virtually all risk of destabilizing the recorded population. A further benefit of this surgery-free procedure is avoiding analgesia and sedation and allowing for an unprecedented, minutes-long turn-around between pre- and post-lesion data collection; the experiment can resume as soon as close observation of the animal is complete, the lesion device is disconnected, and the recording stream is reconnected. The electrolytic lesioning technique is repeatable, because the same multielectrode probe can be used many times to create lesions while maintaining stable electrophysiological recordings. Performing electrolytic lesioning requires passing a specified amount of current through two electrodes (one acting as the anode, the other as the cathode). Selecting which electrodes in the microelectrode array should be the anode and cathode sets the location of the lesion origin. By using a Utah multielectrode array as in this work, the location of the lesion can be changed at a 400µm resolution set by the electrode spacing, while altering the duration and amount of current passed through two electrodes creates changes in the lesion’s spatial extent. Thus, while there is still variation in the precise geometry of damage from each lesion, multielectrode-based lesioning does well to satisfy the second design consideration and is a notable refinement from previous lesioning studies. While this platform was designed for use in rhesus macaques, it could be used in other animals in which multielectrode probes can be implanted, such as other primates, large mammals, and rodents. This technique can also be used in effectively any area of cortex in which a multielectrode probe could be implanted. Therefore, this platform is compatible across many cortical areas, across several species, and theoretically across multielectrode probes, fulfilling the third design consideration. As the lesioning device is a small, low-cost external system that only needs to be connected for the duration of lesioning, it is easy to adopt into existing electrophysiology recording settings. This platform is anticipated to facilitate causal explorations of the relationship between brain and behavior through its unique combination of inactivation and neuroelectrophysiology. Further, it should enable studies of natural reorganization at the neuronal population level.

Results

Electrolytic Lesioning Device and Testing

We created a custom current source circuit which allows us to use the same microelectrode array used to record neuronal activity to also create an electrolytic lesion (Fig. 1). The circuit design is based on simple feedback control circuits TI (2019); Carter and Brown (2016); TI (2020) using an operational amplifier Mancini and Carter (2009) to maintain a constant current. While this technique should theoretically work with any multielectrode probe, we performed all proof-of-concept experiments with a Utah microelectrode array (Blackrock Neurotech, Salt Lake City, UT). Lesion size is controlled by the amplitude and duration of the current passed through the two chosen electrodes of the microelectrode array, although the exact extent of damage will vary for each lesion due to anatomical variation. Our circuit supplies the voltage needed to damage cortex while maintaining high precision in the delivered current (precision error not exceeding 10µA). To find a combination of the current amplitude and duration parameters that would create a reasonable lesion size for our use in rhesus macaques, testing was first performed in ex vivo sheep and pig brains, and then in vivo with anesthetized pigs. Although scientific use of this lesioning device is ultimately intended in rhesus, pilot studies were performed in sheep and pigs in accordance with the guidelines of the National Research Council National Research Council (US) Committee on Guidelines for the Use of Animals in Neuroscience and Behavioral Research (2003); National Research Council (US) Committee for the Update of the Guide for the Care and Use of Laboratory Animals (2011) of replacement, refinement, and reduction (3 R’s). Cortex was lesioned using the chosen parameters, after which histology was collected to provide an understanding of lesion quality and extent. Our exploration was not exhaustive and not designed to fully characterize lesion extent as a function of current amplitude and duration; while being mindful of animal use, a key aim was to converge on a parameter set that yielded small, focal lesions suitable for behavioral studies in rhesus macaques with chronic implants. We both sampled the search space of parameter values, current amplitude and duration, and repeated parameter selections for confirmation. Across eleven ex vivo and five in vivo animals, 61 lesions were performed, including some repeated parameter combinations as biological replicates.

The circuit diagram for the electrolytic lesioning device. An op-amp is used in a negative feedback loop to maintain a constant current through the two electrodes in the brain (RL). The op-amp was implemented as suggested by its accompanying evaluation kit and supplied components. The system is powered by a 12V power supply, and a boost converter is used to create a VCC and -VCC of 450V and -450V, respectively. The current through RL can be set by changing the resistance of the potentiometer, RS. ZP is a hypothesized physiological parasitic component, which could be either resistive or capacitive (dashed box).

Ex Vivo Ovine and Porcine Testing

Initial testing was performed in unfixed cortex from sheep and pigs. Prior studies demonstrated that 20 minutes of 400µA direct current resulted in a spherical cavitation in cortex of approximately 2mm in diameter Wurtz and Goldberg (1972). To create a more localized lesion, parameters were reduced substantially to 250µA direct current for 10 minutes, passed through two adjacent electrodes of a Utah array implanted in ex vivo sheep cortex. This created a clean spheroidal cavitation in cortex approximately 1.5mm in diameter (Fig. 2a,b), which is consistent with lesion sizes that cause measurable behavioral deficits in the motor system Nudo et al. (2003); Wurtz and Goldberg (1972); Glees and Cole (1950). Although lesions of this size are known to cause significant behavioral deficits, we hypothesized that smaller lesions might lead to deficits that were still noticeable but not detrimental to the animal’s overall mobility and mental health, a noted concern in past studies Glees and Cole (1950). Reducing the duration and intensity of the current used to lesion created smaller cavitations in cortex: one minute of 180µA direct current, passed through two adjacent electrodes in ex vivo pig cortex, created a smaller spheroidal cavitation of approximately 0.5mm in diameter (Fig. 2c).

a) Ex vivo demonstration of the electrolytic lesion technique in unfixed sheep cerebral cortex using an intracortical Utah microelectrode array. Sustained delivery of 250µA of direct current for 10 minutes between adjacent electrodes (400µm spacing) resulted in a clean spheroidal cavitation in cortex approximately 1.5mm in diameter. Ruler is marked every 500µm. b) Hematoxylin and eosin (H&E) stained slice of the lesion in (a) clearly shows the lesioned region. Arrows indicate tissue fold artifacts that resulted from the histology process, not the lesion. The other dark pink areas surrounding the cavitation in cortex are regions of necrosis. c) A smaller ex vivo lesion in unfixed cerebral cortex of another sheep created by decreasing the direct current amplitude and duration to 180µA for one minute. The cavitation has a diameter slightly over 0.5mm.

In Vivo Porcine Testing

In vivo testing allowed for further parameter refinement in the presence of natural acute inflammatory responses. By further reducing the duration and intensity of the current used to lesion, the extent of the cavitation in the cortex was reduced to the point where there is no longer a cavitation but instead a clear region of damaged parenchyma. One minute of 150µA direct current, passed through two adjacent electrodes in pig cortex resulted in a well-demarcated region of parenchymal damage (Fig. 3a). The region of parenchymal damage appears paler on H&E staining than the adjacent, unaffected parenchyma and was confined to an upside-down conical region, which was 3.5mm wide at the cortical surface and extended approximately 2mm deep. Within this region (seen magnified in Fig. 3b), there was widespread coagulative necrosis, parenchymal rarefaction, and perivascular microhemorrhage. Acidophilic neuronal necrosis was identified along the marginal boundaries of the region of coagulative necrosis. Adjacent neural parenchyma was unaffected (representative selection shown in Fig. 3c), demonstrating the relatively precise boundary of damage caused by the electrolytic lesion. The physical damage visible as tears in the tissue (white) near the surface of this damaged cone of tissue may be due to withdrawal of the microelectrode array from cortex after testing. In control tests where a microelectrode array was inserted and removed but no current was used to create a lesion (Table S1 and Table S2), regional coagulative necrosis was not present, and cortical damage was confined to mild subcortical and/or perivascular microhemmorage and scattered individual neuronal necrosis, typically in regions adjacent to microvascular hemorrhage. This emphasizes that electrolytic lesioning, not array insertion alone, leads to coagulative necrosis. This testing was performed across many regions of porcine cortex, demonstrating that the electrolytic lesioning technique functions in any area of cortex in which a multielectrode probe can be implanted.

a) H&E stained slice from an in vivo demonstration of the lesioning technique in pig cerebral cortex. 150µA direct current passed through two adjacent electrodes (400µm spacing) for one minute resulted in a conical region of damaged parenchyma. The top of the conical region shows a line of damage which may be caused by physical removal of the microelectrode array after testing. Anatomically observed alterations are clearly demarcated, emphasizing the fine localization of the lesioning method. b) Region of intermixed necrotic and histologically normal neurons within the conical zone of damage is visible in a close-up of the slice from (a). Necrotic neurons have shrunken cell bodies. The microelectrode array is expected to continue recording from remaining healthy neurons after performing a lesion. c) Region of viable neurons outside the conical region of damage is visible in a close-up of the slice from (a). This shows the precise spread of the method, with intact, viable tissue present just outside the lesioned area.

In Vivo Use in Rhesus Macaques

After this validation and refinement, one proof-of-concept lesion was performed in an in vivo sedated rhesus macaque (Monkey F) in order to validate the safety of the procedure. This technique has since been used successfully in experimental settings with two other awake-behaving rhesus macaques (Monkeys H & U), for a total of fourteen lesions using thirteen unique electrode pairs.

Current and Voltage Output

In all of the fourteen lesions across two awake-behaving rhesus macaques, the current source worked as expected, providing a constant current throughout the duration of the procedure. Fluctuations in current amplitude were likely due to the 10µA precision of the multimeter used to read out the current (e.g., the readout would sometimes switch between 150µA and 160µA). The voltage across the microelectrode array also generally behaved as expected (Fig. 4). Upon turning on the lesioning device, the voltage initially increased sharply, sometimes exceeding the slew rate of the voltmeter (seen as a discontinuity in the traces in Fig. 4). Due to the limited resolution of the voltmeter, the voltages were unknown between 0.13 and 0.33s but could not have exceeded 900V based on VCC and -VCC. After this peak, the voltage predominantly levels off.

Voltage traces from seven representative lesions in an awake-behaving rhesus macaque (Monkey H). Lesions are shown in chronological order and are labeled with an experimental ID in the form SYYMMDD, where S indicates the animal, followed by the date. Discontinuity at the beginning of the traces indicates transient voltages that were too rapid to be captured by the voltmeter, lasting between 0.13 and 0.33s. Traces only capture the voltage while the lesioning device was turned on (45 seconds for most lesions and 50 seconds for lesion H200120).

Duration of the applied current is controlled by a switch that cuts the power to the boost converter. However, after the power to the boost converter is removed, the current supplied by the circuit briefly persists — likely due to residual energy present in the capacitors of the boost converter downstream of the switch. In the lesions for which this data was collected, the current persisted at the calibrated intensity for between 2.5 and 4 seconds.

Before performing a lesion, the amount of current to be output by was calibrated using a 50Ω resistor in place of the implanted microelectrode array and altering the resistance level of the external potentiometer. Even when calibrated to generate 150µA, the actual current output when lesioning differed slightly (10 to 20µA above or below the set value). In nine out of the fourteen lesions performed in two awake rhesus macaques (thirteen of which had this voltage data collected), the current value was higher than what was calibrated. We hypothesize that this is due to a parasitic parallel resistance to ground through the animal itself (see dashed box in Fig. 1). To have created the 10 to 20µA increase in current above the set value, this parasitic parallel resistance would have been between 1.2MΩ and 0.6MΩ, respectively, which is in the range of the expected resistance when the body is in dry contact with the environment Fish and Geddes (2009). In the three lesions where the current value was 10 to 20µA lower than what was calibrated, there may have been some parasitic capacitance present. As these parasitic resistances and capacitances arise only when the animal is part of the experimental setup, and they change across sessions, they cannot readily be calculated and accounted for a priori. These parasitics appear small enough not to significantly impact the desired lesion characteristics.

Recording Quality

One advantage of microelectrode arrays is their ability to record from a stable population of cortical neurons over months Vaidya et al. (2014); Dickey et al. (2009); Fraser and Schwartz (2012); Ganguly and Carmena (2009). While the exact neurons captured from the local population vary day-to-day, they largely remain the same, which enables researchers to sample the activity of a consistent neuronal population over time.

Comparisons of the recorded action potential waveforms before and after multiple lesions revealed that microelectrode arrays were able to continue recording stable neuronal activity in awake-behaving rhesus (e.g., lesion six, performed in Monkey H, and lesion 11, performed in Monkey U; Fig. 5a). Surprisingly, even the lesion electrodes themselves continued to record reliably, suggesting that the modest electrolytic lesion intensity used is not prohibitively destructive to an electrode’s recording ability. Given the majority of waveforms appear unaltered (Fig. 5a), it seems unlikely that any acute damage response globally shifted the array away from the recorded neuron population.

a) A representative comparison of recorded action potential waveforms, before and after the sixth and eleventh lesions (top, Monkey H; bottom, Monkey U). The location of the lesion electrodes on the arrays’ spatial layout are marked by black dots. Good quality signal was observed in the recording sessions immediately before and after lesioning (left, right). An action potential detection rate was determined from periods of task engagement (gray-scale shading, capped at 15Hz for visualization). b) As a proxy for neuron loss, a relative turnover in daily recorded neurons was determined by pairwise comparisons of action potential waveforms within three groups: pre-lesion days (pre-pre), pre-lesion versus post-lesion days (pre-post; up to three days post-lesion), and post-lesion days (post-post; four to nine days after a lesion). An inter-quartile range is shaded in grey for the set of 10 consecutive days leading up to lesion 11, the first in Monkey U. The median from this healthy control group was subtracted from all other pairwise comparisons to directly quantify a change in turnover. This pre-lesion group was then combined with the other pre-pre comparisons and tested against the pre-post and post-post groups. The change in the percentage of matching neurons dropped significantly after a lesion (Median test; * < 0.003, ** < 0.0003, corresponding to corrected significance levels of 0.01 and 0.001 for the three comparisons). Note, lesions 11-14 in Monkey U were well-spaced out over three months and considered as independent samples.

After an electrolytic lesion, a small percentage of the recorded waveforms appeared to change significantly, which may be a result of neuron damage or death, or may be an adaptive neuronal circuit response to lesioning. Since the lesion experiments performed in Monkeys H and U were not terminal studies, definitive histological evidence of neuron loss could not be acquired. Therefore, as a proxy for neuron loss, the relative change in the daily turnover of recorded neurons was assessed (Fig. 5b; see Methods for details). To minimize any potential ambiguity, the analysis was restricted to four consecutive lesions in Monkey U (lesions 11-14) that were spaced weeks apart, allowing ample time for stable recovery and for each lesion to be treated as an independent sample — recovery time courses in Monkey H were longer and less cleanly separated.

A healthy control set of the ten days preceding the first lesion (L11) was used to draw pairwise recordings and calculate the median and quartile values (25th and 75th percentiles Fig. 5b). For each lesion, pairwise daily recording sessions were drawn from a set of thirteen contiguous recording days (four pre-lesion days and nine post-lesion days). The pre-post group included up to three days following a lesion, while the post group considered post-lesion days four to nine and was intended to reflect a late post period (not necessarily a complete recovery of brain and behavior). All pairwise comparisons were separated by no more than four days to keep turnover rates stable Gallego et al. (2020). Action potential waveform data from lesions 11-14 and the pre-lesion control set were pooled and grouped into the same pre-pre, post-pre, and post-post categories for a summary comparison (Fig. 5b). Following lesions, the percentage of matching neurons in the pre-post group dropped below the expected percentage of matching neurons among pre-lesion days, appearing to trend back toward comparable levels in the post-lesion group despite still being significantly decreased. This drop in the percentage of matching neurons after lesioning suggests, despite us not having access to histology to confirm, that the lesions terminated neurons.

Discussion

In this report, we demonstrated a novel method for electrolytic lesioning through a microelectrode array that is compatible with electrophysiological recording of neuronal activity pre- and post-lesion. To achieve this, a custom current source was built that would ensure stable current delivery throughout the lesion, for repeated lesions, as well as across different electrode types and animals. This degree of control represents a significant refinement of lesion studies in systems neuroscience research. This model was tested ex vivo with porcine and ovine brains and in vivo with porcine brains. Altering the amount and duration of the current changed the size of the lesion, as evidenced by the histology. Selection of two adjacent lesion electrodes allows for spatial localization under the 16mm2 array. Following these preliminary tests, one lesion was performed in a sedated rhesus macaque to verify the safety of the procedure and obtain primate histology, and fourteen lesions were performed between two awake rhesus macaques. Readouts from the lesioning device itself (both current and voltage) and electrophysiology from the microelectrode array verify that the device delivers the desired power and does not damage the array’s recording ability. We believe that this electrolytic lesioning technique will improve understanding of the motor system by coupling lesioning (an established termination technique) with the detailed spatiotemporal measurements of intracortical electrophysiology Shenoy and Kao (2021); Vaidya et al. (2019).

Although it shares elements with other widely used techniques, performing electrolytic lesioning through a microelectrode array is not a common inactivation technique. For example, injecting current into the brain through microelectrodes is commonly done in intracortical microstimulation Weiss et al. (2019). However, typical current values for microstimulation are around three times smaller than for lesioning, pulse durations are on the order of tens of microseconds, and the pulses by design do not lead to cell death or materially affect the electrodes. Similarly, passing current through microelectrodes to mark their location in cortex is a well-established method for tracking the location of electrodes at the end of an animal study. When histology is eventually performed, the lesion is used to find the anatomical location where electrophysiology was conducted, and it is compatible with single or multi-channel electrodes Chen et al. (2009). However, the intent is different, as electrolytic lesions for marking are not generally used as an inactivation technique. Electrolytic lesioning for inactivation was historically done through a single barrel electrode as the anode and without any post-lesion electrophysiology recordings Wurtz and Goldberg (1972). Therefore, the electrolytic lesioning method for inactivation presented here both is a refinement in precision and is unique in its compatibility with electrophysiology.

Electrolytic lesioning leads to cell death through heat Nikfarjam et al. (2005), electroporation Batista Napotnik et al. (2021), and local changes in pH Nikfarjam et al. (2005). Demonstration of actual neuronal loss in a primate injury model must work within the guidelines of ethical animal research. Therefore, we have triangulated evidence from three distinct sources that indicate the electrolytic lesioning device reliably delivered current that caused tissue damage and neuron death (termination): ex vivo and in vivo histology (Figs. 2, 3), lesion device readouts (Fig. 4), and relative changes in the turnover rate of recorded neurons in vivo (Fig. 5). Despite observing temporary increases in the day-to-day turnover after lesioning (decrease in the percent of matching neurons), it should be noted that this is likely an underestimate as significant neuronal loss may also be occurring nearby, just outside of the electrodes’ recording volume and thus evading detection. The histology presented here indicates regions of partially damaged tissue in which dead neurons are randomly interspersed and whose termination will inevitably vary depending on their proximity to the nearest electrode. It is evident in our recordings that many of the array’s electrodes are recording smaller inseparable waveforms corresponding to multiple neurons located between between 50µm and 140µm from the electrode tip before the lesion Buzsáki (2004); Gerstein and Clark (1964). Even if the closest, more cleanly separated neurons were terminated by the lesion, the recorded signal would still be comprised of the remaining, surviving neurons in that vicinity. These neurons may also become more active after injury resulting in seemingly similar firing rates. This was likely the case for lesion six (Monkey H; Fig. 5a top) and lesion 11 (Monkey U; Fig. 5a bottom), where the average waveforms on the two lesion electrodes themselves appear similar with only slight differences in amplitude. These small shifts will affect the cluster shape when spike-sorting, but are unlikely to be classified as a distinct neuron cluster. Therefore, turnover rate estimates from our in vivo lesions in non-human primates (Fig. 5) are likely underestimating the true extent of neuronal loss. However, there is the potential to overestimate loss as well, given it is impossible to disambiguate between neurons that were silenced but surviving versus truly terminated. Under the constraints, we note that these sources of error may be somewhat counterbalanced.

While we have thoroughly tested this electrolytic lesioning device to ensure proper function, we emphasize that this is an initial design of the device and a preliminary examination of lesion parameters. We have only begun to explore the full potential for electrolytic lesioning through a microelectrode array, in both its technical implementation and its future use in systems neuroscience. This technique was tested across three species, settling on a combination of current amplitude and duration that seemed reasonable for experiments in awake-behaving rhesus macaques. Thorough testing had to be balanced with care and respect for appropriate animal use in accordance with the 3 R’s of the guidelines of the National Research Council National Research Council (US) Committee on Guidelines for the Use of Animals in Neuroscience and Behavioral Research (2003); National Research Council (US) Committee for the Update of the Guide for the Care and Use of Laboratory Animals (2011). Future alterations could be made to the design of the lesioning device itself to optimize lesioning through this platform. For example, electronic control of current delivery for lesioning could be refined through the use of digital switching and a resistor shunt to handle any leakage current. Even with such automation, it would be prudent to maintain a physical switch to prevent extremely long periods of current delivery in the case of a malfunction. A circuit revision with these refinements has been designed and validated, but it has not yet been tested in animals out of aforementioned respect for animal use.

Further exploration into the amplitude and duration parameters of the current used to lesion could lead to markedly different impact on brain tissue. Although these lesions were created with DC current, a variety of current patterns could be used to create a lesion. Catheter ablation of the heart used to be performed with DC current Scheinman (1982); Gallagher et al. (1982). Now, the standard of care is using a radiofrequency (350-500 kHz) alternating current Shivkumar (2019); Morady (1999). Similarly, some have explored using radiofrequency (55kHz) alternating current to create lesions to mark the location of acute electrophysiological recordings Brozoski et al. (2006). It is likely that alternating current or more complex current patterns would lead to different spatial distributions of electrolytic lesions. The PA97 op-amp used in this lesioning device has a slew rate of 8V/µs, and can support AC frequencies in the tens to hundreds of kHz range for the lesioning voltages seen here. At the same time, AC currents would likely introduce new reactive parasitics, potentially along the electrode wire bundle or at interconnects, and should be evaluated carefully.

Different selection of electrodes within the microelectrode array could also lead to different lesion characteristics. We chose to lesion with adjacent electrodes only, in order to create the most focused lesion. In theory, larger ablations are possible by supplying larger currents or performing a lesion over non-adjacent electrodes, which can span as much as 5.6mm for electrodes on the opposite corners of the 4mm × 4mm Utah array.

Electrode shape could also be used to create different lesions. In deep brain stimulation treatment for Parkinson’s disease, new electrodes were designed to directionally focus the stimulation Steigerwald et al. (2019). Similar design could be used in creating intracortical electrodes for elctrophysiology and electrolytic lesioning.

The electrolytic lesioning method presented here enables small, permanent inactivation volumes while maintaining reliable neuroelectrophysiological recordings. Selection of the electrical current pattern, amplitude, and duration, as well as the specific lesion electrodes’ shape and location, offers many different combinations for complementary investigations into the causal role of cortical activity.

Materials and Methods

Electrolytic Lesioning Device Design

In order to control lesion size, a current source is required to stabilize the output, due to changes in local tissue resistivity and the design of intracortical electrodes. Single unit electrodes are generally coated with some type of dielectric material to maintain a low surface area of exposed metal, yielding a high impedance. This high impedance enables the detection of the weak currents (order tens of nA) associated with single neuron action potentials with voltages on the scale of tens to hundreds of microvolts Carter and Shieh (2015). The shafts of the Utah electrode array are insulated with parylene-C (Blackrock Neurotech, Salt Lake City, UT). A small area at the tip of each electrode remains uncoated, through which the ionic current from nearby neurons can be measured. During electrolytic lesioning, applied voltages can be large enough to etch off the dielectric coating from the shaft of the two electrodes used to lesion. As this coating is etched, the impedance of those electrodes falls. Using a constant voltage source to lesion could deliver an inappropriately large amount of current into the brain once the dielectric coating was etched off and impedance reduced, resulting in uncontrolled tissue damage. A constant current source is robust to this changing impedance, maintaining the desired electrical current.

Single unit electrodes are commonly manufactured to have an impedance on the order of 100kΩ to 1MΩ Maynard et al. (1997). In past electrolytic lesioning studies, the currents used to create lesions with corresponding behavioral deficits were on the order of 100µA Glees and Cole (1950); Wurtz and Goldberg (1972). Therefore, in order to supply this current in the face of the high impedance dielectric coating associated with single unit electrodes, high tens to low hundreds of volts are needed, depending on the lesioning parameters chosen. Even though commercial devices exist for electrolytic lesioning (e.g., Ugo Basile, Gemonio, IT), the power supplied may not be sufficient to recreate some current amplitudes used to lesion in literature Wurtz and Goldberg (1972) and generally only support DC waveforms.

We designed and built a custom current source circuit (Fig. 1) that supplies this voltage while maintaining precision in the delivered current (precision error not exceeding 10µA). The circuit features a commercial power operational amplifier, PA97 (Apex Microtechnology, Tucson, AZ), in a negative feedback loop DePaola (2020), implemented in accordance with the corresponding evaluation kit, EK28 (Apex Microtechnology, Tucson, AZ). Power is provided through a 12V external power supply, and a boost converter is used to create a VCC and -VCC of 450V and -450V, respectively. These are both low-cost, off-the-shelf, readily-available discrete circuits. The amplitude of the current supplied through between the cathode and anode to the brain (shown as the load, RL) is tuned through altering the value of a variable potentiometer (RS). The voltage supplied to the op-amp can also be altered by altering the bias resistance with the potentiometer, RB. The total cost of the parts needed is approximately three hundred US dollars, making it an affordable addition to a neuroelectrophysiology recording setup.

Experimental Setup

The lesioning device is built as a contained box, to which the power supply, measurement devices, and microelectrode array are externally connected (Fig. S1). An external switch on the lesioning device controls the power to the boost converter. This is the switch that is manually turned on and off to control the duration of current delivery for lesioning. The lesioning device is connected to two electrodes from the intracortically implanted microelectrode array (RL). All lesioning was performed using Utah electrode arrays with the same specifications (Blackrock Neurotech, Salt Lake City, UT). The arrays are 4mmx4mm, with 96 channels. Electrode shafts are made of silicon (Si) with a metallic outer layer (platinum or platinum-iridium), and coated with parylene-C. Electrodes are 1mm in length and have a 400µm inter-electrode spacing. The array is connected to the lesioning device with the array’s external CerePort (Blackrock Neurotech, Salt Lake City, UT) and a CerePort breakout adaptor that connects to specific electrodes from the array. A Fluke 179 True-RMS Digital Multimeter (Fluke Corporation, Everett, WA) is used as an ammeter in series before RL to measure the current passing through RL (the desired output). A voltmeter spans RL to measure the voltage delivered. The variable potentiometer RS is tuned by a dial on the outside of the lesioning device, allowing quick and easy calibration of the amplitude of the current.

All animal procedures and protocols were approved by the Stanford University Institutional Animal Care and Use Committee (IACUC).

Ex Vivo Ovine and Porcine Testing

Initial testing of lesion parameters was performed ex vivo in three sheep and eight pig brains (Table S1). These un-fixed brains were from animals that were euthanized earlier that day for unrelated research and/or teaching efforts, in line with the reduction principle of animal research National Research Council (US) Committee on Guidelines for the Use of Animals in Neuroscience and Behavioral Research (2003); National Research Council (US) Committee for the Update of the Guide for the Care and Use of Laboratory Animals (2011). As these brains were from unrelated research efforts, we were not given the sex or exact age of the animals. The brain was kept moist with saline throughout the procedure. A Utah array was implanted with a pneumatic insterter (Blackrock Neurotech, Salt Lake City, UT) into suitably large and flat gyri. The lesioning device was then connected to the microelectrode array and used to create a lesion. The location of the array implantation was marked using a histology pen or ink injection and detailed for the research record. The brains were then fixed, sliced, and prepared with hematoxylin and eosin (H&E) staining.

In Vivo Anesthetized Porcine Testing

We next sought to determine how lesion size might be altered in vivo, where blood flow, microhemorrhage, and the body’s acute inflammatory responses could affect the lesion extent and intensity over time. Testing was performed on five anesthetized pigs, to evaluate the ex vivo current amplitude and duration parameters in the context of inflammatory responses (Table S2).

Each pig was sedated and then intubated, ventilated, and placed on inhaled isolflurane. Once under anesthesia and the airway was confirmed, the head was secured in preparation for the procedure. After preparing and cleaning the surgical area, a midline skin incision was made and all skin, muscle, aponeurosis, and periosteum were retracted to expose the skull. Most of the dorsal surface of the skull was exposed along with some of the lateral margins to visualize the anatomy. A single, large craniectomy, exposing most of the superior surface of the dura, was made using a high speed bone drill (ANSPACH, DePuy Synthes, Raynham, MA). Once the craniectomy was at the desired size and location, the dura was washed with saline and hemisphere-wide dural flaps were made to expose the brain. Once exposed, the brain was covered with gauze and kept moist with saline throughout the procedure.

The implantation and lesion procedures were completed in the same manner as in the ex vivo testing. After lesioning, the array was either removed by hand and placed at a new cortical location for further testing or, if the last lesion of the procedure, was left in place for two to three hours before euthanizing the animal to even better mimic the post-lesion inflammatory response. Therefore, histology for all lesions included the physiological response to lesioning after at least two to three hours, and some lesions also retained the array in place as a foreign body during this post-lesion time frame. After the final lesioning was complete and sufficient time had elapsed for lesions to appear histologically, the pig was euthanized via IV injection of beauthanasia solution (100mg/kg, IV). The brain was then fixed, and slices were prepared with H&E staining.

In Vivo Lesions in Rhesus Macaque

Electrolytic lesioning is not painful as there are no direct pain receptors in the central nervous system Hall and Hall (2021). Based on this, and a lack of physiological signs of pain from the anaesthetized pig studies, a lesion was performed on a sedated rhesus macaque (Monkey F; 16 year-old adult, male rhesus macaque) in order to further verify safety before use in awake-behaving rhesus. Again, no physiological signs of stress or complications were observed, increasing confidence for lesions in awake animals. Subsequently, lesions were safely performed in two other rhesus (Monkeys H and U; 14 and 11 years-old, respectively, adult, male rhesus macaques) while awake and seated in a primate chair, without the animals exhibiting any behavioral signs of stress or pain. Across these two animals, fourteen lesions were performed using thirteen unique electrode pairs, demonstrating that the same microelectrode array can be used for several lesions and even the same electrode pair can be used as the anode and cathode for multiple lesions. The Utah array chronically implanted in the monkey’s cortex is connected to the lesioning device via its CerePort (the same one used for recording the neuronal signals from the microelectrode array). The lesioning device is then turned on for a fixed duration, delivering the desired current. The animal is continuously monitored during the procedure for any signs of discomfort. After lesioning is complete and the recording cables are re-connected to the CerePort, the monkey is immediately ready to resume electrophysiological recordings and participate in a wide variety of head-fixed or freely moving behavioral paradigms. Although lesioning itself is painless, the technique is intended to cause a temporary functional impairment. In light of this, monitoring of animal behavior post-procedure is conducted by research staff in coordination with veterinary staff to ensure health, safety, and psychological well-being.

Neuronal Recordings and Processing

Neuronal data was collected at 30kHz from primary motor cortex of Monkeys H and U using a Utah array (described above), captured using a Cerebus system (Blackrock Neurotech, Salt Lake City, UT), then high-pass filtered offline with a fourth order, zero-phase Butterworth filter using a 250 Hz cutoff frequency. Action potentials were identified by thresholding the filtered membrane potential at −4× the root mean square voltage measured over the first minute of the recording session. Data snippets for comparing electrophysiology pre- and post-lesion were generated by taking the preceding 16 sampled data points (0.53ms), and proceeding 32 sampled points (1.07ms) around these identified threshold crossings. Waveforms whose peak-to-trough amplitude exceeded 300µV are atypical of extracellular recordings from cortical neurons and were excluded Chestek et al. (2011). Waveform width was calculated as the difference between the two prominent peaks in the temporal derivative of the extracellular potential. Unphysiological waveforms whose widths were longer than 1ms were also excluded Bean (2007). The remaining identified action potential waveforms were then averaged for each electrode and event rates were determined as the number of spikes on a given electrode divided by the duration of the recording session (excluding brief periods of inactivity before the start of the session).

As a putative indication of neuronal loss, the relative change in neuron turnover was measured between daily recording sessions (each pair separated by at most three days). For a given electrode, identified action potentials from the first day’s recording session were spike-sorted, then the same waveforms were pooled with the second day’s waveforms and sorted again. Sorting was performed by projecting a given set of waveforms onto their top two principal components and running hierarchical clustering on these scatter points using the fclusterdata function from SciPy’s cluster.hierarchy package (SciPy version 1.6.3) Virtanen et al. (2020). The distance threshold for cluster identification was set as 4.5 (arbitrary units) as it appeared to represent the data well and generated results consistent with turnover rates observed in healthy primate cortex Gallego et al. (2020). Although a range of threshold values would likely be valid, the precise value is not a concern here since we are seeking to identify relative changes in neuron turnover. If fewer clusters were identified when waveforms from both days were combined (neurons lost) or more clusters were identified (new neurons detected), then that electrode was designated as altered by the lesion. It is assumed that previously undetected neurons largely reflect compensation to the loss of other nearby neurons, which were not recorded by the array or were present in the background multi-unit activity. The fact that the majority of waveforms on the array are consistently the same alleviates the concern that simple shifts in the relative position of the tissue explains the altered waveforms. Otherwise, electrodes with matching waveforms were deemed unaltered and presented as a percentage of the array’s total 96 electrodes for that pair of days (% match). While inter-spike interval statistics are commonly used as a feature to help identify distinct neurons in healthy animals, they are likely not appropriate here, since surviving neurons may drastically alter their activity after injury.

These pairwise clustering comparisons were then split up into three groups: pre-lesion versus pre-lesion days (pre-pre), pre-lesion versus post-lesion days (pre-post), and post-lesion version post-lesion days (post-post). Although a set of thirteen consecutive days was analyzed for a given lesion (four pre- and nine post-), only recording session pairs that were separated by four-days or fewer were analyzed in order to minimize the potential confound of increased turnover resulting from more than a few days of separation Gallego et al. (2020). Comparisons were then made between the pre-lesion period, an acute injury period (up to three days following a lesion), and a late recovery period (days four to nine following a lesion). The Median test was used to test the null hypothesis that lesioning had no effect on recorded turnover with 99% and 99.9% confidence intervals Bonferroni corrected for the three statistical comparisons made between groups (±/3, significant if p < 0.003, * or p < 0.0003, **). The ten pre-lesion days leading up to the first lesion in Monkey U were also compared to establish a median and inter-quartile range as a healthy control. This median value, 72%, was subtracted from the percentage change in each group to determine a relative change in the matching waveforms between days after injury (Δ% match).

eLife Assessment

This paper reports a valuable new method for creating localized damage to candidate brain regions for functional and behavioral studies. The authors present solid support for their ability to create long-term local lesions with mm spatial resolution, although data supporting the extension to primates was incomplete. If the authors can demonstrate that the technique is applicable beyond the specific technical setup they employ, the paper is likely to be of broad interest to brain researchers working to establish causal links between neural circuits and behavior.

Reviewer #1 (Public Review):

In the paper, the authors illustrated a novel method for Electrolytic Lesioning through a microelectronics array. This novel lesioning technique is able to perform long-term micro-scale local lesions with a fine spatial resolution (mm). In addition, it allows a direct comparison of population neural activity patterns before and after the lesions using electrophysiology. This new technique addresses a recent challenge in the field and provides a precious opportunity to study the natural reorganization/recovery at the neuronal population level after long-term lesions. It will help discover new causal insights investigating the neural circuits controlling behavior.

Several minor concerns are summarized below:

It was not always clear what the lesion size was. This information is important for future applications, for example, in the visual cortex, where neurons are organized in retinotopy patterns.

It would be helpful if the author could add some discussion about whether and how this method could be used in other types of array/multi-contact electrodes, such as passive neuropixels, S-probes, and so on. In addition, though an op-amp was used in the design, it would still be helpful if the author could provide a recommended range for the impedance of the electrodes.

Reviewer #2 (Public Review):

This work by Bray et al. presented a customized way to induce small electrolytic lesions in the brain using chronically implanted intracortical multielectrode arrays. This type of lesioning technique has the benefit of high spatial precision and low surgical complexity while allowing simultaneous electrophysiology recording before, during, and after the lesion induction. The authors have validated this lesioning method with a Utah array, both ex vivo and in vivo using pig models and awake-behaving rhesus macaques. Given its precision in controlling the lesion size, location, and compatibility with multiple animal models and cortical areas, the authors believe this method can be used to study cortical circuits in the presence of targeted neuronal inactivation or injury and to establish causal relationships before behavior and cortical activity.

Strengths:

Great presentation of design considerations that addressed the gaps of current lesioning and neuronal inactivation methods, especially the cross-compatibility that allows this method to be used across different cortical regions and in different animal models, making it easy to be adopted into a variety of electrophysiology studies.

This method can induce lesions that are highly precise and repeatable in size and location, allowing for robust investigation of neuronal circuit function. When combined with the ability to record without disruption both in the acute and chronic phase after lesioning, this would create a great tool to study neural adaptation and reorganization.

The customized current source is simple, low-cost, yet effective in delivering precise, controllable current for electrolytic lesioning, and thus easy to adopt for a range of neuroscience applications.

Extensive ex vivo testing and validation were performed before moving into in vivo and eventually nonhuman primate (NHP) experiments, successfully reducing animal use.

Weaknesses:

In many of the figures, it is not clear what is shown and the analysis techniques are not well described.

The flexibility of lesioning/termination location is limited to the implantation site of the multielectrode array, and thus less flexible compared to some of the other termination methods outlined in Appendix 2.

Although the extent of the damage created through the Utah array will vary based on anatomical structures, it is unclear what is the range of lesion volumes that can be created with this method, given a parameter set. It was also mentioned that they performed a non-exhaustive parameter search for the applied current amplitude and duration (Table S1/S2) to generate the most suitable lesion size but did not present the resulting lesion sizes from these parameter sets listed. Moreover, there's a lack of histological data suggesting that the lesion size is precise and repeatable given the same current duration/amplitude, at the same location.

It is unclear what type of behavioral deficits can result from an electrolytic lesion this size and type (~3 mm in diameter) in rhesus macaques, as the extent of the neuronal loss within the damaged parenchyma can be different from past lesioning studies.

The lesioning procedure was performed in Monkey F while sedated, but no data was presented for Monkey F in terms of lesioning parameters, lesion size, recorded electrophysiology, histological, or behavioral outcomes. It is also unclear if Monkey F was in a terminal study.

As an inactivation method, the electrophysiology recording in Figure 5 only showed a change in pairwise comparisons of clustered action potential waveforms at each electrode (%match) but not a direct measure of neuronal pre and post-lesioning. More evidence is needed to suggest robust neuronal inactivation or termination in rhesus macaques after electrolytic lesioning. Some examples of this can be showing the number of spike clusters identified each day, as well as analyzing local field potential and multi-unit activity.

The advantages over recently developed lesioning techniques are not clear and are not discussed.

There is a lack of quantitative histological analysis of the change in neuronal morphology and loss.

There is a lack of histology data across animals and on the reliability of their lesioning techniques across animals and experiments.

There is a lack of data on changes in cortical layers and structures across the lesioning and non-lesioning electrodes.

Author response:

The following is the authors’ response to the current reviews.

We thank the reviewers and editor for their careful review of our work. We believe the resulting manuscript is much stronger. We agree with the comments made by Reviewer #2 regarding additional histology and neuronal data analysis, which will be presented in subsequent work.

The following is the authors’ response to the original reviews.

Reviewer 1 (Public Weaknesses):

It was not always clear what the lesion size was. This information is important for future applica- tions, for example, in the visual cortex, where neurons are organized in retinotopy patterns.

We thank the reviewer for this feedback. While there is some variation in lesion volume for a given parameter set, we have added more details of the volumes of lesions created in our testing (Fig. 4 and Fig. 5).

It would be helpful if the author could add some discussion about whether and how this method could be used in other types of array/multi-contact electrodes, such as passive neuropixels, S- probes, and so on. In addition, though an op-amp was used in the design, it would still be helpful if the author could provide a recommended range for the impedance of the electrodes.

We thank the reviewer for this suggestion. We have both added a demonstration of use in a differ- ent multielectrode probe type (with a U-probe) in Fig. 8, and we have added a discussion about which types of multielectrode probes would be suitable on Page 15, Line 420.

“We demonstrated that our electrolytic lesioning technique works with a linear multicontact probe by testing with a U-Probe in ex vivo rabbit cortex. There are no particular limitations that would prevent our specific electrolytic lesioning technique and device from working with any passive multielectrode probe. The main requirements for use are that the probe has two electrodes that can directly (via whatever necessary adapters) connect to the lesioning device, such that arbitrary current can be passed into them as the anode and cathode. This would limit use of probes, like Neuropixels, where the on-chip acquisition and digitization circuitry generally precludes direct connection to electrodes [1], [2]. The impedance of the multielectrode probe should not be an issue, due to the use of an op amp. We showed use with a Utah array (20-800 kΩ) and a U-Probe (1-1.5 MΩ). The specific op amp used here has a voltage range of ± 450 V, which assuming a desired output of 150 µA of current would limit electrode impedance to 6 MΩ. Though a different op amp could easily be used to accommodate a higher electrode impedance, it is unlikely that this would be necessary, since most electrodes have impedances between 100 kΩ to 1 MΩ [3].”

Reviewer 2 (Public Weaknesses):

In many of the figures, it is not clear what is shown and the analysis techniques are not well described.

We thank the reviewer for this feedback. We hope that our edits to both the figures and the text have improved clarity for readers.

The flexibility of lesioning/termination location is limited to the implantation site of the multielec- trode array, and thus less flexible compared to some of the other termination methods outlined in Appendix 2.

We thank the reviewer for this point. You are right that the lesioning location is limited to the multielectrode array’s implantation site, while other methods in Appendix 2 do not require prox- imity of the lesion location and the electrophysiology recording site. However, we believe that the closeness of the lesioning location to the microelectrode array is a strength - guaranteeing record- ings from the perilesional area - even with the small negative of reduced flexibility. Multielectrode arrays can be implanted in many areas of cortex. If one wanted to study distal effects of a lesion, additional electrophysiology probes could be implanted to record from those areas. We have noted this on Page 3, Line 117.

“While the link between the lesion location and the multielectrode location technically con- strains the lesion to an area of cortex in which a multielectrode array could be implanted, we see the connection as a positive, because it ensures recording some neuroelectrophysiology from the perilesional area in which recovery is hypothesized to occur (see Appendix 1Data Availabilityappendix.41).”

Although the extent of the damage created through the Utah array will vary based on anatomical structures, it is unclear what is the range of lesion volumes that can be created with this method, given a parameter set. It was also mentioned that they performed a non-exhaustive parameter search for the applied current amplitude and duration (Table S1/S2) to generate the most suitable lesion size but did not present the resulting lesion sizes from these parameter sets listed. Moreover, there’s a lack of histological data suggesting that the lesion size is precise and repeatable given the same current duration/amplitude, at the same location.

We thank the reviewer for this thoughtful feedback. We have added figures (Figs. 4 and 5), where we show the relationship between estimated lesion volume and the current amplitude and duration parameters. These figures include more data from the tests in Supplementary File 1 and Supplementary File 2. While there is some variation in lesion volume for a given current amplitude and duration, there is still a clear relationship between the parameters and lesion volume.

It is unclear what type of behavioral deficits can result from an electrolytic lesion this size and type (∼3 mm in diameter) in rhesus macaques, as the extent of the neuronal loss within the damaged parenchyma can be different from past lesioning studies.

While we appreciate the reviewer’s interest in the behavioral deficits associated with our lesions in rhesus macaques, reporting these falls beyond the scope of this manuscript. Future work will explore the behavioral deficits associated with these lesions

The lesioning procedure was performed in Monkey F while sedated, but no data was presented for Monkey F in terms of lesioning parameters, lesion size, recorded electrophysiology, histological, or behavioral outcomes. It is also unclear if Monkey F was in a terminal study.

We apologize for not being more explicit about the parameters used for the lesion in Monkey F. We have added this in Results on Page 5, Line 209 and in Methods on Page 19, Line 586.

“After this validation and refinement, one proof-of-concept lesion (150 µA direct current passed through adjacent electrodes for 45 seconds) was performed in an in vivo sedated rhe- sus macaque (Monkey F) in order to validate the safety of the procedure.”

“This lesion was created by applying 150 µA of direct current to two adjacent electrodes in the microelectrode array for 45 seconds.”

We also clarified the parameters used for the other lesions in Monkeys H and U in Results on Page 7, Line 233 and in Methods on Page 19, Line 586.

“In all of the fourteen lesions across two awake-behaving rhesus macaques (150 µA direct current passed through adjacent electrodes for 30 or 45 seconds (30s for Monkey U and 45s for Monkey H, except lesion H200120 which was for 50 seconds)), the current source worked as expected, providing a constant current throughout the duration of the procedure.”

“In these lesions, 150 µA of direct current was applied to two adjacent electrodes in the mi- croelectrode array for 30 or 45 seconds (30s for Monkey U, 45s for Monkey H), except in lesion H200120 where current was applied for 50 seconds.”

Monkey F was euthanized shortly after the lesion, so we now mention this on Page 19, Line 583.

“Based on this, and a lack of physiological signs of pain from the anaesthetized pig studies, a lesion was performed on a sedated rhesus macaque who was subsequently euthanized due to unrelated health complications (Monkey F; 16 year-old adult, male rhesus macaque) in order to further verify safety before use in awake-behaving rhesus.”

Because Monkey F was sedated and then euthanized shortly after, there is no behavioral data. As the lesion in sedated Monkey F was used to validate the safety of the procedure, any further data and analysis fall beyond the scope of this manuscript.

As an inactivation method, the electrophysiology recording in Figure 5 only showed a change in pairwise comparisons of clustered action potential waveforms at each electrode (%match) but not a direct measure of neuronal pre and post-lesioning. More evidence is needed to suggest robust neuronal inactivation or termination in rhesus macaques after electrolytic lesioning. Some exam- ples of this can be showing the number of spike clusters identified each day, as well as analyzing local field potential and multi-unit activity.

The reviewer has pointed out some short comings of the original analysis, which we believe have since been addressed with the revised analysis. LFP and spiking activity are functional measures that are more ambiguous in terms of loss and are also the subject of another manuscript currently under revision.

The advantages over recently developed lesioning techniques are not clear and are not discussed.

We thank the reviewer for noting this. We have added a section, also responding to their later request for us to compare our work to Khateeb et al. 2022, by adding a section to the Discussion on Page 16, Line 434.

“Perhaps the most unique advantage of our technique in comparison with other existing inactivation methods lies in Design Consideration #1: stable electrophysiology pre- and post-inactivation (Appendix 1Data Availabilityappendix.41). While several methods exist that allow for localization and size control of the inactivation (Design Consideration #2) and cross compatibility across regions and species (Design Consideration #3), few have achieved compatibility with stable electrophysiology. For example, some studies record electrophysiology only after the creation of the lesion, preventing comparison with baseline neuronal activity [4]. One recent study, Khateeb, et al., 2022, developed an inactivation method that is effectively combined with stable electrophysiology by creating photothrombotic lesions through a chronic cranial window integrated with an electrocorticography (ECoG) array [5], which may be appropriate for applications where local field potential (LFP) recording is sufficient. This approach has trade-offs with regards to the three design considerations presented in Appendix 1Data Availabilityappendix.41.

While Khateeb, et al., present a toolbox with integrated, stable electrophysiology from an ECoG array pre- and post- inactivation (Design Consideration #1), it demonstrated recordings from an ECoG array with limited spatial resolution. While a higher density ECoG array that would provide higher spatial resolution could be used, increasing the density of opaque electrodes might occlude optical penetration and constrain photothrombotic lesions. Further, ECoG arrays are limited to recording LFP, not electrophysiology at single neuron resolution, potentially missing meaningful changes in the neuronal population activity after lesioning. Khateeb, et al., demonstrated localization and control the size of inactivation (Design Consideration #2). In this manuscript, we have shown that the amount and duration of direct current are significant determinants of lesion size and shape, while with photothrombotic lesions, light intensity and aperture diameter are the significantly relevant parameters. One potential advantage of photothrombotic approaches is the use of optical tools to monitor anatomical and physiological changes after lesioning through the cranial window, though the research utility of this monitoring remains to be demonstrated.

Although the method presented by Khateeb, et al., shows some cross-compatibility (Design Consideration #3), it has greater limitations in comparison with the method presented here. For example, while Khateeb, et al., notes that the approach could be adapted for use in smaller organisms, no modification is needed for use in other species with this work’s approach–so long as a multielectrode probe is implantable. In this manuscript we demon- strate electrolytic lesioning spanning two multielectrode probes across rabbits, pigs, sheep, and rhesus macaques, and our same device could be easily used with other smaller species, like rats, in which multielectrode probes have been successfully implanted [6]. Further, the approach in Khateeb, et al., is limited to superficial brain structures, due to the need for opti- cal accessibility. As noted, fiber optics could allow access to deeper structures, which would bring associated additional tissue damage, but deeper structure lesioning was not demon- strated. In contrast, the approach presented here can be used in any region of cortex in which a multielectrode probe can be implanted, which, depending on the probe used, does not limit it to surface structures. For example, we demonstrated use of our lesioning tech- nique with a linear U-probe (Fig. 8figure.caption.25), which could be used to reach deeper layers of cortex or specific deep cortical structures. In both techniques, the location of the lesion is tied to the location of the electrophysiology (for Khateeb et al., wherever the cra- nial window and ECoG array are; for this technique, wherever the multielectrode probe has been implanted), which ensures that the electrophysiology will include recordings from the perilesional area. Neither work addresses the potential of their technique to induce chronic post-lesion behavioral effects, which is a key goal for future work.”

There is a lack of quantitative histological analysis of the change in neuronal morphology and loss.

We appreciate the reviewer’s desire for a quantitative histological analysis, however this falls out- side of the scope of this manuscript. We are not attempting to make strong claims about the number of neurons lost through lesioning or thoroughly characterize morphological changes in the neurons. The histology is intended to show that lesioning did lead to a loss of neurons, but the precise num- ber of neurons lost is neither in scope nor is likely to be highly conserved across lesions.

There is a lack of histology data across animals and on the reliability of their lesioning techniques across animals and experiments.

We thank the reviewer for this point. As stated above, we have now added Fig. 4 and Fig. 5, which includes volume estimates based on the histology from more of our ex vivo and in vivo testing across animals.

There is a lack of data on changes in cortical layers and structures across the lesioning and non- lesioning electrodes.

We acknowledge that the histology does not have the level of detail that is expected from many modern studies. However, the goal here was dramatically different: we sought to calibrate a novel lesion device, ensure it’s safe use in large mammals (specifically, non-human primates) and pro- vide estimates of the lesion size to compare with the literature. The extent of histology that could be performed and the tools available to us prevent such an in depth analysis. We can say based on shank length of the Utah arrays used and known anatomy that we have affected layer 2/3 and maybe a bit of layer 4.

Reviewer 1 (Recommendations For The Authors):

Figure 5b. It would be helpful if the author could plot the delta match separately for the lesion elec- trodes, near neighbor electrodes, and far neighbors. This would help understand the lesion effect, specifically whether the effect is selective (e.g., more potent for the lesion and adjacent electrodes.)

The fact that neuron loss is not particularly selective can already be seen in the spike waveform plots, arranged spatially on the array. Plenty of clear change is observed far from the lesion elec- trodes (marked with black dots) as well as nearby. We have made mention of this localized non- specificity in the main text and have ensured to remphasize in the figure legened. While a nice suggestion, we currently don’t feel this result rises to the level of a figure given it is not highly specific spatially.

Reviewer 2 (Recommendations For The Authors):

Overall the quality of the paper, the figures and the analysis used could be significantly improved. There is a lack of scientific rigor in the presentation of figures and analysis techniques. It is not clear what the authors are trying to communicate through the figures and their choice of figures to show is confusing (see below).

We thank the reviewer for their pointed critiques and believe we have addressed their concerns with many changes to the text, a revamped waveforms analysis, and both the expansion and addition of results.

The neurophysiology data shown doesn’t suggest neuronal loss, it only shows change which needs strong control data to show it is due to a lesion.

As detailed below, we have presented a revised analysis that provides this control. While the reviewer is right to point out we can distinguish actual neuron loss from neuron silencing, we be- lieve the new analysis rigorously indicates new rates of sample turnover beyond those expected from healthy state.

The histology figure should be replaced with a high-quality representation without folds.

We understand the reviewer’s suggestion. While ideally we would have many histology slices from each lesion, due to cost, we were only able to collect one histology slice per lesion. The folds were introduced by the company that performed the H&E staining, and we unfortunately cannot remove the folds. Therefore, despite the folds, this is the best and only image from this lesion. We hope that the markings on the figure and the comment in the caption is sufficient to explain to readers that the folds are not a result of the lesion but instead a result of the histology process.

The authors suggest that this lesioning method will be compatible with any available multielec- trode probe theoretically. Since all testing was done with a Utah array, it will be helpful to add an explanation about potential constraints that will make a given array compatible with this method.

We thank the reviewer for this suggestion. As stated above, we have both added a demonstration of use in a different multielectrode probe type (with a U-probe) in Fig. 8, and we have added a discussion about which types of multielectrode probes would be suitable on Page 15, Line 420.

The authors should cite and discuss previous studies using electrolytic lesioning in awake-behaving animals to study the causal connection between the brain and behavior. (One example study: Morissette MC, Boye SM. Electrolytic lesions of the habenula attenuate brain stimulation reward. Behavioural brain research. 2008 Feb 11;187(1):17-26.)

We thank the reviewers for this suggestion. We have added a mention of existing electrolytic le- sioning studies on Page 2, Line 88.

“Prior termination studies mostly measure behavioral output, with no simultaneous measures of neuronal activity during the behavior, impairing their ability to provide insight into the causal connection between the brain and behavior [7]–[11], or with no baseline (i.e., pre- lesion) measures of neuronal activity [4].”

The authors should compare their technique with other recent lesioning studies in primates (e.g. Khateeb et al, 2022)

We again thank the reviewer for this point. Specifically not mentioning Khateeb et al. 2022 was a submission error on our part; we cited the paper in Appendix 2 in the version uploaded to the eLife submission portal, but we had uploaded the version prior to citing it to bioRxiv. We have combined addressing this with addressing a previous comment, as mentioned above, with a section in the Discussion on Page 16, Line 434.

In Appendix 2, the authors suggest that a major limitation of optogenetics and chemogenetic in- activation methods is the lack of rhesus-compatible constructs. However, several viral constructs have successful implementation in rhesus monkeys so far (e.g. Galvan A, Stauffer WR, Acker L, El-Shamayleh Y, Inoue KI, Ohayon S, Schmid MC. Nonhuman primate optogenetics: recent advances and future directions. Journal of Neuroscience. 2017 Nov 8;37(45):10894-903; Tremblay et al, Neuron 2020)

We thank the reviewer for pointing us to these papers. We have added a more thorough description of what we meant by lack of rhesus-compatible constructs in that Appendix.

“However, other challenges exist with using optogenetics as an inactivation method in nonhu- man primates, including difficulty reliably affecting behavior [12]. While several constructs for rhesus macaques have been developed [13], [14], reports of successfully inducing be- havioral effects have a small effect size and are less numerous than might be expected [12], and several null results have been published [15]–[17]. Other remaining challenges include the need to develop a head-mounted, battery powered light delivery system for multi-day delivery of light and difficulty integrating illumination with simultaneous chronic neuro- electrophysiology.”

For Figure 5b, only pairwise comparison results from monkey U (L11-14) are shown. It is unclear why such results from monkey H were shown in Figure 5a but not in 5b.

We thank the reviewer for pointing out this unconventional one monkey result. As described in the original submission, we previously omitted Monkey H from the analysis in Figure 5b (now Figure 7) since some of the lesions were closely spaced together, preventing well defined pre- and post- lesion rates of turnover. Never-the-less we have included Monkey H in all the revised analysis and believe even the less cleanly separated data shows useful indications of neuron loss or silencing evoked by the lesion.

Behavioral data (during a motor task) from the awake behaving monkeys (U and H) would greatly strengthen the claim that this lesioning method is capable of creating a behavioral effect and can be adopted to study the relationship between neural function and behavior outcomes.

While we are grateful for the reviewer’s interest in the application of our lesioning technique to studies involving behavior, a behavioral analysis of the effects of our electrolytic lesions falls be- yond the scope of this Tools and Resources manuscript. We would also like to point out that we do not claim that we have achieved a behavioral deficit in this manuscript.

Figure 2 would benefit from an illustration of the Utah array placement and the location of the sites used for lesioning. The authors can either overlay the illustrations on the current ex-vivo and histology images or create a separate schematic to demonstrate that for the readers. Also, Figure 2B needs to be replaced with one without the folds to avoid confusion for the readers.

We have added Figure 2 - figure supplement 1, which shows both the location within the Utah array of the two electrodes used to create the lesions as well as the relative size of the surface area of the lesion and the array. Unfortunately, as the lesion was created under the array, the exact location of the array relative to the lesion is unknown.

As mentioned above, Figure 2B is the only histological image from that lesion. We hope that the markings in the image as well as the caption sufficiently explain that the folds are unrelated to the lesion itself.

Figure 3, the conical region is not well delineated. Data across animals and lesion volume with respect to different parameters should be included.

We have included a supplemental figure, Figure 3 - figure supplement 1, where we have used a dashed white line to clearly indicate the area of damaged parenchyma, in case it was not clear in Figure 3a. We have also added volume estimates from lesions across animals and different param- eters. The ex vivo estimates are shown in Figure 4 and the in vivo estimates are shown in Figure 5.

Figure 4: it is not clear what is being communicated, and where the voltage traces are from.

We thank the reviewer for noting this confusion. We have added some lines in the text to explain what the voltage traces show, both in the caption to Fig. 6 and in the text on Page 7, Line 238.

“Traces only capture the values while the lesioning device was turned on (45 seconds for most lesions and 50 seconds for lesion H200120). A) Voltage traces. Discontinuity at the beginning of the traces indicates transient voltages that were too rapid to be captured by the voltmeter, lasting between 0.13 and 0.33 s. The fluctuating voltages, especially the rapid in- crease in voltage at the beginning of lesioning, emphasize the importance of using a current source to deliver consistent amounts of current into the brain.”

“The voltage across the microelectrode array fluctuated much more than the current did, em- phasizing that we made the correct choice in using a current source to ensure delivery of consistent amounts of current into the brain (Fig. 6figure.caption.19).”

Figure 5: why did the authors choose to use matching units as a measure of the lesion? It is surprising that there are still units on the location that the authors claim to be a lesion. To clarify that it would be helpful to show the location of the lesion in Figure 4a. Also, what can we conclude about the lesion induction when we see units on the lesion electrode? The change in unit match shows that there is a change in the network (although the authors need to show control for that so we know those changes don’t happen due to natural dynamics). It is not clear what is the time duration for pre-pre and post-post (i.e. minutes, seconds, hours). Do these comparisons come from the same time frame or are they coming from two fragments of time for both pre and post- conditions?

Aside from post-mortem histology and tissue assays, there is no good way to confirm neuron loss with chronically implanted electrode arrays in nonhuman primates. Waveforms were chosen as they are the one readily isolated physical measure of the system we are injuring. Although functional measures of activity could indicate neuron loss (topic of following papers), there are many conceivable changes in firing rate patterns that could manifest spuriously as loss, making the estimation of loss even more ambiguous and challenging this way.

We believe the new Figure 7 will make the procedure much more clear, while also providing the control requested by the reviewer, illustrating that new statistical categories of altered waveforms emerge during a lesion, beyond those associated with typical changes in waveform composition within multi-unit recordings seen during recording sample turnover fom healthy animals. We further note that by confining this analysis to four day spans at most, we have limited the impact of daily sample turnover described in the literature (Gallego, 2020).

The time duration for pre-session versus pre-session (pre-post and post-post), is some multiple of the approximate 24 hours between each daily recording session. Therefore, since restricting our- selves to four days separation, between 24 and 96 hours. Spikes are sampled from successful trial periods (so on the order of seconds, compiled into minutes across the whole recording session). Although already described in the main text, these points have been reemphasized in the figure legend.

CNO (line 931) needs to be explained.

We thank the reviewer for this point. We have defined CNO and its relevance in Appendix 2.

“Additionally, chronic inactivation over days may be logistically challenging, as the half life of clozapine N-oxide (CNO, a ligand used to activate DREADD receptors) is on the order of hours.”

Version 2

Abstract

Abstract

Lesion studies have historically been instrumental for establishing causal connections between brain and behavior. They stand to provide additional insight if integrated with multielectrode techniques common in systems neuroscience. Here we present and test a platform for creating electrolytic lesions through chronically implanted, intracortical multielectrode probes without compromising the ability to acquire neuroelectrophysiology. A custom-built current source provides stable current and allows for controlled, repeatable lesions in awake-behaving animals. Performance of this novel lesioning technique was validated using histology from ex vivo and in vivo testing, current and voltage traces from the device, and measurements of spiking activity before and after lesioning. This electrolytic lesioning method avoids disruptive procedures, provides millimeter precision over the extent and submillimeter precision over the location of the injury, and permits electrophysiological recording of single-unit activity from the remaining neuronal population after lesioning. This technique can be used in many areas of cortex, in several species, and theoretically with any multielectrode probe. The low-cost, external lesioning device can also easily be adopted into an existing electrophysiology recording setup. This technique is expected to enable future causal investigations of the recorded neuronal population’s role in neuronal circuit function, while simultaneously providing new insight into local reorganization after neuron loss.

Introduction

Neuroelectrophysiology – using electrodes to record the electrical signals generated by neurons – has been the defining technique that shifted neuroscience from macroscale anatomy and loss-of-function studies down to the microscale activity of individual neurons and synapses in a given region of interest Sporns (2016). Prior to the last few decades, electrophysiology was limited to simultaneous recordings of at most five neurons with small numbers of electrodes (typically one, two, or four) McNaughton et al. (1983). The development of neuroelectrophysiology recording techniques with a large number of electrodes starting in the 1970s Wise et al. (1970); Campbell et al. (1990, 1991); Rios et al. (2016); Hong and Lieber (2019) was a boon to systems neuroscience, enabling simultaneous recording from hundreds of neurons and revealing previously unseen aspects of population coding Cunningham and Yu (2014); Kalaska (2019); Shenoy and Kao (2021). Over the last five years, electrode count has continued to increase, with new probes containing thousands of electrodes Jun et al. (2017); Steinmetz et al. (2021); Obaid et al. (2020); Sahasrabuddhe et al. (2020); Musk (2019). While experiments and analysis have revealed population activity that correlates strongly with behavioral output Yu et al. (2009); Churchland et al. (2012); Sadtler et al. (2014); Golub et al. (2018); Elsayed et al. (2016); Goldman et al. (2019); Gallego et al. (2020); Liu et al. (2020); Rasmussen et al. (2017), novel tools that both record from and inactivate neurons are required to establish causal connections to behavior Vaidya et al. (2019); Shenoy and Kao (2021); Slonina et al. (2022). Although many inactivation methods exist, to date it has been challenging to find a repeatable, long-term inactivation technique compatible with chronic intracortical neuroelectrophysiology.

Design Considerations and Existing Inactivation Methods

Three main design considerations are required for a technique that successfully combines electro-physiology with inactivation (further detailed in Appendix 1).

Stable electrophysiology pre- and post-inactivation: Avoiding physical disruption enables direct comparison of pre-lesion activity with both the acute and chronic stages of injury.

Ability to localize and control the size of the inactivation: Precise focal inactivations strike a balance between being large enough to alter performance but small enough to spare sufficient tissue to record local re-organization and recovery.

Cross-compatibility: A technique that can be used in many areas of cortex, with any multielectrode probe, and in several species will enable causal investigation across a large variety of contexts. A method that works across-species (especially in large-animals like rhesus macaques) would leverage the existing injury literature, recording technologies, and behavioral assays from rodents to new world monkeys, while also being well-suited for the behavioral sophistication and human homology of macaques Higo (2021).

Neuronal activity can either be temporarily or permanently inactivated, defined as manipulation or termination, respectively Vaidya et al. (2019). For clarity, termination refers to a technique that causes death of neurons, removing them from the circuit. We non-exhaustively review existing manipulation and termination techniques in Appendix 2. However, none of these existing techniques meet all of the above design considerations.

Manipulations enable a causal understanding of the relationship between neuronal activity and behavior by studying adaptation to and from a perturbation Slonina et al. (2022). Existing manipulation methods include intracortical microstimulation, optogenetics, pharmacology, transcranial stimulation, cooling loops, and chemogenetics. Since neurons remain anatomically and physiologically viable, manipulations can be easily repeated, until desensitization occurs.

As termination causes cell death, it can generate stronger causal evidence than a manipulation Vaidya et al. (2019). For example, even though a transient manipulation of neuronal activity may temporarily disrupt behavior, only a sustained manipulation could elicit the system’s longterm adaption Slonina et al. (2022). Over the days Ferrier and Yeo (1884) and potentially weeks Bundy and Nudo (2019); Zeiler et al. (2015) following a termination, the surrounding circuitry and broader network may adapt, leading to behavioral recovery and demonstrating that although the terminated region was causally implicated in behavioral control in the moment, the terminated neurons are not themselves causally necessary. A sustained manipulation could accomplish a similar effect to a termination at the systems level, up to the point at which the manipulation stopped. However, it can be difficult to create sustained inactivation with existing reversible manipulation methods, limiting their ability to study the brain’s natural reorganization over timescales of days to months.

Termination methods overcome the problem of sustained and consistent inactivation from which temporary inactivation techniques suffer. Therefore, they enable a form of causal inference not possible with temporary inactivation methods Vaidya et al. (2019); Shenoy and Kao (2021). Existing termination techniques include mechanical trauma, endovascular occlusion, Rose Bengal mediated photothrombosis, and chemical lesioning. Prior termination studies mostly measure behavioral output, with no simultaneous measures of neuronal activity during the behavior, impairing their ability to provide insight into the causal connection between the brain and behavior Morissette and Boye (2008); Nudo et al. (2003); Nudo (2013); Wurtz and Goldberg (1972); Glees and Cole (1950), or with no baseline (i.e., pre-lesion) measures of neuronal activity Khanna et al. (2021).

Electrolytic Lesioning Through A Microelectrode Array

In order to best meet the three design considerations, we created a device to make an electrolytic lesion through the same microelectrode array used to record neuronal activity. The dual use of this microelectrode array achieves the first design consideration of stable electrophysiology pre- and post-inactivation, as it enables recording from the exact same signal source with minimal physical disruption or displacement. Aside from the initial implantation of the array, there are no invasive procedures required, removing virtually all risk of destabilizing the recorded population. A further benefit of this surgery-free procedure is avoiding analgesia and sedation and allowing for an unprecedented, minutes-long turn-around between pre- and post-lesion data collection; the experiment can resume as soon as close observation of the animal is complete, the lesion device is disconnected, and the recording stream is reconnected. The electrolytic lesioning technique is repeatable, because the same multielectrode probe can be used many times to create lesions while maintaining stable electrophysiological recordings. Performing electrolytic lesioning requires passing a specified amount of current through two electrodes (one acting as the anode, the other as the cathode). Selecting which electrodes in the microelectrode array should be the anode and cathode sets the location of the lesion origin. By using a Utah multielectrode array as in this work, the location of the lesion can be changed at a 400 µm resolution set by the electrode spacing, while altering the duration and amount of current passed through two electrodes creates changes in the lesion’s spatial extent. Thus, while there is still variation in the precise geometry of damage from each lesion, multielectrode-based lesioning does well to satisfy the second design consideration and is a notable refinement from previous lesioning studies. While this platform was designed for use in rhesus macaques, it could be used in other animals in which multielectrode probes can be implanted, such as other primates, large mammals, and rodents. This technique can also be used in effectively any area of cortex in which a multielectrode probe could be implanted. While the link between the lesion location and the multielectrode location technically constrains the lesion to an area of cortex in which a multielectrode array could be implanted, we see the connection as a positive, because it ensures recording some neuroelectrophysiology from the perilesional area in which recovery is hypothesized to occur (see Appendix 1). Therefore, this platform is compatible across many cortical areas, across several species, and theoretically across multielectrode probes, fulfilling the third design consideration. As the lesioning device is a small, low-cost external system that only needs to be connected for the duration of lesioning, it is easy to adopt into existing electrophysiology recording settings. This platform is anticipated to facilitate causal explorations of the relationship between brain and behavior through its unique combination of inactivation and neuroelectrophysiology. Further, it should enable studies of natural reorganization at the neuronal population level.

Results

Electrolytic Lesioning Device and Testing

We created a custom current source circuit which allows us to use the same microelectrode array used to record neuronal activity to also create an electrolytic lesion (Fig. 1). The circuit design is based on simple feedback control circuits Texas Instruments Incorporated (2019); Carter and Brown (2016); Texas Instruments Incorporated (2020) using an operational amplifier to maintain a constant current Carter and Mancini (2009). While this technique should theoretically work with any multielectrode probe, we performed all proof-of-concept experiments with a Utah microelectrode array (Blackrock Neurotech, Salt Lake City, UT). Lesion size is controlled by the amplitude and duration of the current passed through the two chosen electrodes of the microelectrode array, although the exact extent of damage will vary for each lesion due to anatomical variation. Our circuit supplies the voltage needed to damage cortex while maintaining high precision in the delivered current (precision error not exceeding 10 µA). To find a combination of the current amplitude and duration parameters that would create a reasonable lesion size for our use in rhesus macaques, testing was first performed in ex vivo sheep and pig brains, and then in vivo with anesthetized pigs.

The circuit diagram for the electrolytic lesioning device. An op-amp is used in a negative feedback loop to maintain a constant current through the two electrodes in the brain (RL). The op-amp was implemented as suggested by its accompanying evaluation kit and supplied components. The system is powered by a 12 V power supply, and a boost converter is used to create a VCC and -VCC of 450 V and −450 V, respectively. The current through RL can be set by changing the resistance of the potentiometer, RS. ZP is a hypothesized physiological parasitic component, which could be either resistive or capacitive (dashed box).

Although scientific use of this lesioning device is ultimately intended in rhesus, pilot studies were performed in sheep and pigs in accordance with the guidelines of the National Research Council National Research Council (US) Committee on Guidelines for the Use of Animals in Neuroscience and Behavioral Research (2003); National Research Council (US) Committee for the Update of the Guide for the Care and Use of Laboratory Animals (2011) of replacement, refinement, and reduction (3 R’s). Cortex was lesioned using the chosen parameters, after which histology was collected to provide an understanding of lesion quality and extent. Our exploration was not exhaustive and not designed to fully characterize lesion extent as a function of current amplitude and duration; while being mindful of animal use, a key aim was to converge on a parameter set that yielded small, focal lesions suitable for behavioral studies in rhesus macaques with chronic implants. We both sampled the search space of parameter values, current amplitude and duration, and repeated parameter selections for confirmation. Across eleven ex vivo and five in vivo animals, 61 lesions were performed, including some repeated parameter combinations as biological replicates.

Ex Vivo Ovine and Porcine Testing

Initial testing was performed in unfixed cortex from sheep and pigs. Prior studies demonstrated that 20 minutes of 400 µA direct current resulted in a spherical cavitation in cortex of approximately 2 mm in diameter Wurtz and Goldberg (1972). To create a more localized lesion, parameters were reduced substantially to 250 µA direct current for 10 minutes, passed through two adjacent electrodes of a Utah array implanted in ex vivo sheep cortex. This created a clean spheroidal cavitation in cortex approximately 1.5 mm in diameter (Fig. 2A,B and Figure 2 - figure supplement 1A,B), which is consistent with lesion sizes that cause measurable behavioral deficits in the motor system Nudo et al. (2003); Wurtz and Goldberg (1972); Glees and Cole (1950). Although lesions of this size are known to cause significant behavioral deficits, we hypothesized that smaller lesions might lead to deficits that were still noticeable but not detrimental to the animal’s overall mobility and mental health, a noted concern in past studies Glees and Cole (1950). Reducing the duration and intensity of the current used to lesion created smaller cavitations in cortex: one minute of 180 µA direct current, passed through two adjacent electrodes in ex vivo pig cortex, created a smaller spheroidal cavitation of approximately 0.5 mm in diameter (Fig. 2C and Figure 2 - figure supplement 1C,D).

A) Ex vivo demonstration of the electrolytic lesion technique in unfixed sheep cerebral cortex using an intracortical Utah microelectrode array. Sustained delivery of 250 µA of direct current for 10 minutes between adjacent electrodes (400 µm spacing) resulted in a clean spheroidal cavitation in cortex approximately 1.5 mm in diameter. Ruler is marked every 500 µm. B) Hematoxylin and eosin (H&E) stained slice of the lesion in (A) clearly shows the lesioned region. Arrows indicate tissue fold artifacts that resulted from the histology process, not the lesion. The other dark pink areas surrounding the cavitation in cortex are regions of necrosis. C) A smaller ex vivo lesion in unfixed cerebral cortex of a pig created by decreasing the direct current amplitude and duration to 180 µA for one minute. The cavitation has a diameter slightly over 0.5 mm.

In Vivo Porcine Testing

In vivo testing allowed for further parameter refinement in the presence of natural acute inflammatory responses. By further reducing the duration and intensity of the current used to lesion, the extent of the cavitation in the cortex was reduced to the point where there is no longer a cavitation but instead a clear region of damaged parenchyma. One minute of 150 µA direct current, passed through two adjacent electrodes in pig cortex resulted in a well-demarcated region of parenchymal damage (Fig. 3A and Figure 3 - figure supplement 1). The region of parenchymal damage appears paler on H&E staining than the adjacent, unaffected parenchyma and was confined to an upside-down conical region, which was 3.5 mm wide at the cortical surface and extended approximately 2 mm deep. Within this region (seen magnified in Fig. 3B), there was widespread coagulative necrosis, parenchymal rarefaction, and perivascular microhemorrhage. Acidophilic neuronal necrosis was identified along the marginal boundaries of the region of coagulative necrosis. Adjacent neural parenchyma was unaffected (representative selection shown in Fig. 3C), demonstrating the relatively precise boundary of damage caused by the electrolytic lesion. The physical damage visible as tears in the tissue (white) near the surface of this damaged cone of tissue may be due to withdrawal of the microelectrode array from cortex after testing. In control tests where a microelectrode array was inserted and removed but no current was used to create a lesion (Supplementary File 1and Supplementary File 2), regional coagulative necrosis was not present, and cortical damage was confined to mild subcortical and/or perivascular microhemmorhage and scattered individual neuronal necrosis, typically in regions adjacent to microvascular hemorrhage. This emphasizes that electrolytic lesioning, not array insertion alone, leads to coagulative necrosis. This testing was performed across many regions of porcine cortex, demonstrating that the electrolytic lesioning technique functions in any area of cortex in which a multielectrode probe can be implanted.

A) H&E stained slice from an in vivo demonstration of the lesioning technique in pig cerebral cortex. 150 µA direct current passed through two adjacent electrodes (400 µm spacing) for one minute resulted in a conical region of damaged parenchyma. The top of the conical region shows a line of damage which may be caused by physical removal of the microelectrode array after testing. Anatomically observed alterations are clearly demarcated, emphasizing the fine localization of the lesioning method. Note: An outline of the region of damaged parenchyma is shown in Figure 3 - figure supplement 1. B) Region of intermixed necrotic and histologically normal neurons within the conical zone of damage is visible in a close-up of the slice from (A). Necrotic neurons have shrunken cell bodies. The microelectrode array is expected to continue recording from remaining healthy neurons after performing a lesion. C) Region of viable neurons outside the conical region of damage is visible in a close-up of the slice from (A). This shows the precise spread of the method, with intact, viable tissue present just outside the lesioned area.

Estimated lesion volumes from lesions created in ex vivo sheep and pig cortex. All volumes were cavitations in cortex and were estimated assuming a conical lesion volume. A) Estimated volumes are shown with black dots for each direct current amplitude and duration pairing. B) Estimated volumes are shown for a variety of durations of applied 180 µA direct current. A curve was fit to this line, showing an exponential relationship between duration of the current and lesion volume (V = (5.4 × 10−7) × e5.6t, R2 = 0.413).

Estimated lesion volumes from lesions created in in vivo pig cortex for a subset of direct current amplitude and duration pairings. For rarefied tissue damage, the lesion volume is indicated with a black dot, while for cavitation damage, the volume is indicated with a blue dot. Estimated lesion volumes were calculated from histology measurements and assumed a conical lesion volume (with the histology slice bisecting the volume), except for the lesion where 150 µA direct current was applied for 30 s, which was calculated as a spherical volume as indicated by the visually identified damage in the histology slice.

In Vivo Use in Rhesus Macaques

After this validation and refinement, one proof-of-concept lesion (150 µA direct current passed through adjacent electrodes for 45 seconds) was performed in an in vivo sedated rhesus macaque (Monkey F) in order to validate the safety of the procedure. This technique has since been used successfully in experimental settings with two other awake-behaving rhesus macaques (Monkeys H & U), for a total of fourteen lesions using thirteen unique electrode pairs.

Relationship Between Applied Current and Lesion Volume

Our ex vivo ovine and porcine testing and our in vivo porcine testing demonstrated a relationship between the amplitude and duration of the direct current applied and the size of the lesion. In order to characterize this relationship, we compared lesion volume estimates for a variety of direct current amplitudes and durations. We had slightly fewer data points for the in vivo lesions, due to difficulty in obtaining histology slices that captured the damage from the lesion within the slice. Lesion volumes were estimated from histology or photographs taken from our ex vivo and in vivo testing. Volumes were estimated using volume formulas for the approximate geometries observed visually in the histology. Most volumes were estimated assuming a conical lesion volume, chosen based on Figure 2 and Figure 3, though one in vivo volume was visually identified as spherical and estimated as such. We found that as both the amplitude and duration of direct current increase, the volume of the lesion also increases. For ex vivo lesions with a current amplitude of 180 µA, we found that volume was an exponential function of the duration of current (V = (5.4×10−7)×e5.6t, R2 = 0.413). We also found that as amplitude and duration of direct current decrease, the type of damage changes. While lesions with larger current amplitudes and durations create cavitations in cortex (as in all of the ex vivo and the longest in vivo lesion), lesions where the current had a lower amplitude and/or was applied for a shorter duration resulted in rarefied tissue damage (as seen in Fig. 3), containing parenchymal rarefaction, coagulative necrosis, and perivascular microhemorrhage.

Current and Voltage Output

In all of the fourteen lesions across two awake-behaving rhesus macaques (150 µA direct current passed through adjacent electrodes for 30 or 45 seconds (30s for Monkey U and 45s for Monkey H, except lesion H200120 which was for 50 seconds)), the current source worked as expected, providing a constant current throughout the duration of the procedure. Fluctuations in current amplitude were likely due to the 10 µA precision of the multimeter used to read out the current (e.g., the readout would sometimes switch between 150 µA and 160 µA). The voltage across the microelectrode array fluctuated much more than the current did, emphasizing that we made the correct choice in using a current source to ensure delivery of consistent amounts of current into the brain (Fig. 6). Upon turning on the lesioning device, the voltage initially increased sharply, sometimes exceeding the slew rate of the voltmeter (seen as a discontinuity in the traces in Fig. 6). Due to the limited resolution of the voltmeter, the voltages were unknown between 0.13 and 0.33 s but could not have exceeded 900V based on VCC and -VCC. After this peak, the voltage predominantly levels off.

Voltage and current traces from seven representative lesions in an awake-behaving rhesus macaque (Monkey H). Lesions are shown in chronological order and are labeled with an experimental ID in the form SYYMMDD, where S indicates the animal, followed by the date. Traces only capture the values while the lesioning device was turned on (45 seconds for most lesions and 50 seconds for lesion H200120). A) Voltage traces. Discontinuity at the beginning of the traces indicates transient voltages that were too rapid to be captured by the voltmeter, lasting between 0.13 and 0.33 s. The fluctuating voltages, especially the rapid increase in voltage at the beginning of lesioning, emphasize the importance of using a current source to deliver consistent amounts of current into the brain. B) Current traces. The device delivered stable current for the duration of the lesion. Fluctuations are likely due to the 10 µA precision of the multimeter that read the current.

Duration of the applied current is controlled by a switch that cuts the power to the boost converter. However, after the power to the boost converter is removed, the current supplied by the circuit briefly persists — likely due to residual energy present in the capacitors of the boost converter downstream of the switch. In the lesions for which this data was collected, the current persisted at the calibrated intensity for between 2.5 and 4 seconds.

Before performing a lesion, the amount of current to be output by was calibrated using a 50 kΩ resistor in place of the implanted microelectrode array and altering the resistance level of the external potentiometer. Even when calibrated to generate 150 µA, the actual current output when lesioning differed slightly (10 to 20 µA above or below the set value). In nine out of the fourteen lesions performed in two awake rhesus macaques (thirteen of which had this voltage data collected), the current value was higher than what was calibrated. We hypothesize that this is due to a parasitic parallel resistance to ground through the animal itself (see dashed box in Fig. 1). To have created the 10 to 20 µA increase in current above the set value, this parasitic parallel resistance would have been between 1.2 MΩ and 0.6 MΩ, respectively, which is in the range of the expected resistance when the body is in dry contact with the environment Fish and Geddes (2009). In the three lesions where the current value was 10 to 20 µA lower than what was calibrated, there may have been some parasitic capacitance present. As these parasitic resistances and capacitances arise only when the animal is part of the experimental setup, and they change across sessions, they cannot readily be calculated and accounted for a priori. These parasitics appear small enough not to significantly impact the desired lesion characteristics.

Recording Quality

One advantage of microelectrode arrays is their ability to record from a stable population of cortical neurons over months Vaidya et al. (2014); Dickey et al. (2009); Fraser and Schwartz (2012); Ganguly and Carmena (2009). While the exact neurons captured from the local population vary day-to-day, they largely remain the same, which enables researchers to sample the activity of a consistent neuronal population over time.

Comparisons of the recorded action potential waveforms before and after multiple lesions revealed that microelectrode arrays were able to continue recording stable neuronal activity in awake-behaving rhesus (e.g., after one lesion performed in Monkey U and after six lesions performed in Monkey H; Fig. 7A). Surprisingly, even the lesion electrodes themselves continued to record reliably, suggesting that the modest electrolytic lesion intensity used is not prohibitively destructive to an electrode’s recording ability. Given the majority of waveforms appear similar (Fig. 7A), it seems unlikely that any acute damage response physically shifted the array away from the recorded neuron population.

A) A representative comparison of recorded action potential waveforms, before and after the first lesion in Monkey U (top) and the sixth lesion in Monkey H (bottom). The location of the lesion electrodes are marked by black dots, showing visible changes in waveforms with low spatial specificity across the array. Signal was observed in the recording sessions immediately before and after lesioning (left, right). An action potential detection rate was determined from both periods of rest and task engagement (gray-scale shading, capped at 5 Hz for visualization). B) The multineuron activity recorded on each electrode was analyzed for changes in the proportions of activity. Action potential waveforms from a selected day (black) were compared to 1000 waveforms from a previous day (grey) by projecting the selected day’s waveforms into the top two principal components determined from the previous day. The median radius of all waveform points relative to the origin in these two dimensions are represented as circles for both the comparison and original day. The difference in radii, Δr, was computed for each channel and for all pairs of recording days, separated by no more than four days to ensure minimal rates of spontaneous turnover noted in the literature Gallego et al. (2020). Changes in radius could arise through changes in the relative proportion of activity among the recorded neurons, as well as putative neuron loss (r ≪ 0) and gain (r ≫ 0). C) The day-to-day pairwise comparisons fall within three groups: pre-lesion days (pre-pre), pre-lesion versus post-lesion days (pre-post; up to three days post-lesion), and post-lesion days (post-post; four to seven days after a lesion). In all comparison cases, 24, 48, 72, or 96 hours separated the recording sessions. Distribution of the Δr values for all channels and days are shown for each group from Monkey U. Central panels Gaussian mixture models were fit to the data and the optimal number of components for each was determined using the Bayesian information criterion (BIC, normalized arbitrary units). An entirely new cluster is identified for the pre-post group that is hypothesized to largely represent the loss of neurons from the local population, beyond the usual rate of turnover observed in pre-lesion conditions. Right panel The three groups were then compared by looking at the percentage of the 96 electrodes that matched across comparison days. The percentage of matching neurons dropped significantly after a lesion (median test; * < 0.017, corresponding to the conservative Bonferroni corrected significance threshold of 0.05 for the three comparisons). D) The same analysis is performed for Monkey H, which yielded two cases: lesions consistent with Monkey U (lesions 2, 4, and 7; left) and those with high levels of turnover before and after injury (lesions 5, 6, 8, 9, and 10; right). Note, lesions in Monkey U were well-spaced out over three months and considered as independent samples. Lesions in Monkey H were performed in much quicker succession, which likely contributes to the discrepancy.

After an electrolytic lesion, a fraction of the recorded waveforms appeared to change significantly, which may be a result of neuron damage or death, or may be an adaptive neuronal circuit response to lesioning. Since the lesion experiments performed in Monkeys H and U were not terminal studies, definitive histological evidence of neuron loss could not be acquired. Therefore, as a proxy for neuron loss, the relative change in the daily turnover of recorded neurons was assessed (Fig. 7B; see Methods for details).

For each lesion, pairwise daily recording sessions were drawn from a set of thirteen contiguous recording days (four pre-lesion days and nine post-lesion days). For each electrode, principal component analysis was performed on standard normalized, sampled waveforms from the first of the two days; next, waveforms sampled from the second day were normalized and projected onto the top two dimensions determined from the first day 7B). A circle that captured half of the scatter points measures the spread, and changes in its radius were assumed to reflect changes in the recorded multi-unit activity. Typically, the difference in radius was nearly zero, reflecting similar proportions of activity from a similar composition of nearby neurons. Modest decreases and increases in radius were also frequently observed and the projected waveforms appeared similar. Two final categories were observed less commonly, in which the radius decreased or increased dramatically and the projected waveforms were substantially different.

Comparison days were grouped as pre-pre, pre-post, which included an acute period of up to three days following a lesion, and post-post, which compared post-lesion days four to seven and was intended to reflect a late post period (not necessarily a complete recovery of brain and behavior). All pairwise comparisons were separated by no more than four days to minimize natural turnover rates described in the literature Gallego et al. (2020). Following lesions in Monkey U (7C) and Monkey H (case 1; 7D), a distinct cluster was detected in the histogram of radial changes (Δr) for all array electrode comparisons between pre-lesion and acute post-lesion days. This group captures the either complete silencing of neurons or their putative loss and appeared (significantly) in the pre-post group only. As a measure of turnover across the array, a % match was determined for each groups’ pairwise comparison between daily recording sessions, which reflects the number of electrodes where Δr remained unchanged. These values dropped below the expected percentage of matching neurons among pre- and post-lesion days (pre-pre, post-post) confirming that rates of turnover in the recorded population had been accelerated by the lesion. Although similar results were obtained for case 2 in Monkey H, they were not significant (7D). This is largely due much larger variability in turnover and the fact that these lesions in Monkey H were performed before complete recovery was observed.

Use of Electrolytic Lesioning Device with Other Multielectrode Probes

In order to verify that this method for creating electrolytic lesions through a multielectrode probe works with different probe types, we performed tests using a Plexon U-Probe - a linear electrode with 24 contacts spaced 100 µm apart - in ex vivo rabbit cortex. Using our original electrolytic lesioning device and the U-Probe, we created five distinct lesions through contiguous contacts (100 µm apart) by applying differing current amplitudes (240 µA and 150 µA) for differing durations (30, 60, and 120 s). In Figure 8, we show the location of the five lesions, as well as the current and voltage traces (as in Figure 6). This testing demonstrates the ability to deliver steady current through a different multielectrode probe from the Utah microelectrode array, which was the primary probe used in most of these experiments.

Ex vivo testing in rabbit cortex using a linear multielectrode probe. Surface penetration locations for the five lesions made with a linear multielectrode probe are marked on rabbit cortex, with corresponding voltage traces (A) and current traces (B) arranged top-to-bottom. Electrode contacts were separated by 100 µm and lowered approximately 400 µm below the surface. The voltage slowly decreased throughout each lesion, while the circuit maintained steady current delivery. Only minor fluctuations in current of less than 10 µA were observed, based on the resolution of the ammeter.

Discussion

In this report, we demonstrated a novel method for electrolytic lesioning through a microelectrode array that is compatible with electrophysiological recording of neuronal activity pre- and post-lesion. To achieve this, a custom current source was built that would ensure stable current delivery throughout the lesion, for repeated lesions, as well as across different electrode types and animals. This degree of control represents a significant refinement of lesion studies in systems neuroscience research. This model was tested ex vivo with porcine and ovine brains and in vivo with porcine brains. Altering the amount and duration of the current changed the size of the lesion, as evidenced by the histology. Selection of two adjacent lesion electrodes allows for spatial localization under the 16 mm2 array. Following these preliminary tests, one lesion was performed in a sedated rhesus macaque to verify the safety of the procedure, and fourteen lesions were performed between two awake rhesus macaques. Readouts from the lesioning device itself (both current and voltage) and electrophysiology from the microelectrode array verify that the device delivers the desired power and does not damage the array’s recording ability. We believe that this electrolytic lesioning technique will improve understanding of the motor system by coupling lesioning (an established termination technique) with the detailed spatiotemporal measurements of intracortical electrophysiology Shenoy and Kao (2021); Vaidya et al. (2019).

Although it shares elements with other widely used techniques, performing electrolytic lesioning through a microelectrode array is not a common inactivation technique. For example, injecting current into the brain through microelectrodes is commonly done in intracortical microstimulation Weiss et al. (2019). However, typical current values for microstimulation are around three times smaller than for lesioning, pulse durations are on the order of tens of microseconds, and the pulses are charged balanced so they by design do not lead to cell death or materially affect the electrodes. Similarly, passing current through microelectrodes to mark their location in cortex is a well-established method for tracking the location of electrodes at the end of an animal study. When histology is eventually performed, the lesion is used to find the anatomical location where electrophysiology was conducted, and it is compatible with single or multi-channel electrodes Chen et al. (2009). However, the intent is different, as electrolytic lesions for marking are not generally used as an inactivation technique. Electrolytic lesioning for inactivation was historically done through a single barrel electrode as the anode and without any post-lesion electrophysiology recordings Wurtz and Goldberg (1972). Therefore, the novel electrolytic lesioning method for inactivation presented here both is a refinement in precision and is unique in its compatibility with electrophysiology.

Electrolytic lesioning leads to cell death through heat Nikfarjam et al. (2005), electroporation Batista Napotnik et al. (2021), and local changes in pH Nikfarjam et al. (2005). Demonstration of actual neuronal loss in a primate injury model must work within the guidelines of ethical animal research. Therefore, we have triangulated evidence from three distinct sources that indicate the electrolytic lesioning device reliably delivered current that caused tissue damage and neuron death (termination): ex vivo and in vivo histology (Figs. 2, 3), lesion device readouts (Fig. 6), and relative changes in the turnover rate of recorded neurons in vivo (Fig. 7C and D, case 1).

Temporary increases in the day-to-day turnover of recorded waveforms were observed after lesioning (i.e., decreases in the percent of matching neurons). In addition to increases in turnover characteristic of pre-lesion data, we further identified statistically new clusters of lesion evoked perturbations to low-dimensional projections of recorded neurons’ waveforms. On one hand, this is likely an underestimate, as significant neuronal loss is likely occurring just outside of the array electrodes’ recording volume and thus evading detection. However, there is also the potential to overestimate loss as well, given it is impossible to disambiguate between neurons that were silenced but surviving versus truly terminated. We note that these sources of error may be somewhat counterbalanced. The evidence provided here suggests a temporary increase in the recorded waveforms’ turnover after a lesion, which likely captures evoked neuron loss in addition to lesionevoked changes in the proportions of activity recorded among neurons.

It is evident in our recordings that many of the array’s electrodes are recording smaller inseparable waveforms (hash) corresponding to many different neurons located between between 50 µm and 140 µm from the electrode tip before the lesion Buzsáki (2004); Gerstein and Clark (1964). Even if the closest, cleanly separable neurons were terminated by the lesion, the recorded signal would still be comprised of the remaining, surviving neurons in that vicinity. These neurons may also become more active after a lesion, resulting in similar estimated firing rates. This was likely the case for the first lesion in Monkey U (Fig. 7A, top) and lesion six (Monkey H; Fig. 7A bottom), where the average waveforms on the two lesion electrodes themselves appear similar with only slight differences in amplitude.

While we have thoroughly tested this electrolytic lesioning device to ensure proper function, we emphasize that this is an initial design of the device and a preliminary examination of lesion parameters. We have only begun to explore the full potential for electrolytic lesioning through a microelectrode array, in both its technical implementation and its future use in systems neuroscience. This technique was tested across three species, before settling on a combination of current amplitude and duration that seemed reasonable for experiments in awake-behaving rhesus macaques. Thorough testing had to be balanced with care and respect for appropriate animal use in accordance with the 3 R’s of the guidelines of the National Research Council National Research Council (US) Committee on Guidelines for the Use of Animals in Neuroscience and Behavioral Research (2003); National Research Council (US) Committee for the Update of the Guide for the Care and Use of Laboratory Animals (2011). Future alterations could be made to the design of the lesioning device itself to optimize lesioning through this platform. For example, electronic control of current delivery for lesioning could be refined through the use of digital switching and a resistor shunt to handle any leakage current. Even with such automation, it would be prudent to maintain a physical switch to prevent extremely long periods of current delivery in the case of a malfunction. A circuit revision with these refinements has been designed and validated, but it has not yet been tested in animals out of aforementioned respect for animal use.

Further exploration into the amplitude and duration parameters of the current used to lesion could lead to markedly different impact on brain tissue. Although these lesions were created with direct current, a variety of current patterns could be used to create a lesion. Catheter ablation of the heart used to be performed with direct current Scheinman (1982); Gallagher et al. (1982). Now, the standard of care is using a radiofrequency (350-500 kHz) alternating current Shivkumar (2019); Morady (1999). Similarly, some have explored using radiofrequency (55kHz) alternating current to create lesions to mark the location of acute electrophysiological recordings Brozoski et al. (2006). It is likely that alternating current or more complex current patterns would lead to different spatial distributions of electrolytic lesions. The PA97 op-amp used in this lesioning device has a slew rate of 8V∕μs, and can support AC frequencies in the tens to hundreds of kHz range for the lesioning voltages seen here. At the same time, alternating currents would likely introduce new reactive parasitics, potentially along the electrode wire bundle or at interconnects, and should be evaluated carefully.

Different selection of electrodes within the microelectrode array could also lead to different lesion characteristics. We chose to lesion with adjacent electrodes only, in order to create the most focused lesion. In theory, larger ablations are possible by supplying larger currents or performing a lesion over non-adjacent electrodes, which can span as much as 5.6 mm for electrodes on the opposite corners of the 4 mm x 4 mm Utah array.

Electrode shape could also be used to create different lesions. In deep brain stimulation treatment for Parkinson’s disease, new electrodes were designed to directionally focus the stimulation Steigerwald et al. (2019). Similar design could be used in creating intracortical electrodes optimized for both elctrophysiology and electrolytic lesioning. We demonstrated that our electrolytic lesioning technique works with a linear multicontact probe by testing with a U-Probe in ex vivo rabbit cortex. There are no particular limitations that would prevent our specific electrolytic lesioning technique and device from working with any passive multielectrode probe. The main requirements for use are that the probe has two electrodes that can directly (via whatever necessary adapters) connect to the lesioning device, such that arbitrary current can be passed into them as the anode and cathode. This would limit use of probes, like Neuropixels, where the on-chip acquisition and digitization circuitry generally precludes direct connection to electrodes Jun et al. (2017); Steinmetz et al. (2021). The impedance of the multielectrode probe should not be an issue, due to the use of an op amp. We showed use with a Utah array (20-800 kΩ) and a U-Probe (1-1.5 MΩ). The specific op amp used here has a voltage range of ± 450 V, which assuming a desired output of 150 µA of current would limit electrode impedance to 6 MΩ. Though a different op amp could easily be used to accommodate a higher electrode impedance, it is unlikely that this would be necessary, since most electrodes have impedances between 100 kΩ to 1 MΩ Maynard et al. (1997).

Perhaps the most unique advantage of our technique in comparison with other existing inactivation methods lies in Design Consideration #1: stable electrophysiology pre- and post-inactivation (Appendix 1). While several methods exist that allow for localization and size control of the inactivation (Design Consideration #2) and cross compatibility across regions and species (Design Consideration #3), few have achieved compatibility with stable electrophysiology. For example, some studies record electrophysiology only after the creation of the lesion, preventing comparison with baseline neuronal activity Khanna et al. (2021). One recent study, Khateeb, et al., 2022, developed an inactivation method that is effectively combined with stable electrophysiology by creating photothrombotic lesions through a chronic cranial window integrated with an electrocorticography (ECoG) array Khateeb et al. (2022), which may be appropriate for applications where local field potential (LFP) recording is sufficient. This approach has trade-offs with regards to the three design considerations presented in Appendix 1.

While Khateeb, et al., present a toolbox with integrated, stable electrophysiology from an ECoG array pre- and post-inactivation (Design Consideration #1), it demonstrated recordings from an ECoG array with limited spatial resolution. While a higher density ECoG array that would provide higher spatial resolution could be used, increasing the density of opaque electrodes might occlude optical penetration and constrain photothrombotic lesions. Further, ECoG arrays are limited to recording LFP, not electrophysiology at single neuron resolution, potentially missing meaningful changes in the neuronal population activity after lesioning. Khateeb, et al., demonstrated localization and control the size of inactivation (Design Consideration #2). In this manuscript, we have shown that the amount and duration of direct current are significant determinants of lesion size and shape, while with photothrombotic lesions, light intensity and aperture diameter are the significantly relevant parameters. One potential advantage of photothrombotic approaches is the use of optical tools to monitor anatomical and physiological changes after lesioning through the cranial window, though the research utility of this monitoring remains to be demonstrated.

Although the method presented by Khateeb, et al., shows some cross-compatibility (Design Consideration #3), it has greater limitations in comparison with the method presented here. For example, while Khateeb, et al., notes that the approach could be adapted for use in smaller organisms, no modification is needed for use in other species with this work’s approach–so long as a multielectrode probe is implantable. In this manuscript we demonstrate electrolytic lesioning spanning two multielectrode probes across rabbits, pigs, sheep, and rhesus macaques, and our same device could be easily used with other smaller species, like rats, in which multielectrode probes have been successfully implanted Black et al. (2018). Further, the approach in Khateeb, et al., is limited to superficial brain structures, due to the need for optical accessibility. As noted, fiber optics could allow access to deeper structures, which would bring associated additional tissue damage, but deeper structure lesioning was not demonstrated. In contrast, the approach presented here can be used in any region of cortex in which a multielectrode probe can be implanted, which, depending on the probe used, does not limit it to surface structures. For example, we demonstrated use of our lesioning technique with a linear U-probe (Fig. 8), which could be used to reach deeper layers of cortex or specific deep cortical structures. In both techniques, the location of the lesion is tied to the location of the electrophysiology (for Khateeb et al., wherever the cranial window and ECoG array are; for this technique, wherever the multielectrode probe has been implanted), which ensures that the electrophysiology will include recordings from the perilesional area. Neither work addresses the potential of their technique to induce chronic post-lesion behavioral effects, which is a key goal for future work.

The electrolytic lesioning method presented here enables small, permanent inactivation volumes while maintaining reliable neuroelectrophysiological recordings. Selection of the electrical current pattern, amplitude, and duration, as well as the specific lesion electrodes’ shape and location, offers many different combinations for complementary investigations into the causal role of cortical activity. Further, the same microelectrode array used to create the lesion and record electrophysiology can be used for intracortical microstimulation. This additional aspect of our platform will allow for testing of intracortical microstimulation as a possible treatment or therapy after neuronal loss.

Materials and Methods

Electrolytic Lesioning Device Design

In order to control lesion size, a current source is required to stabilize the output, due to changes in local tissue resistivity and the design of intracortical electrodes. Single unit electrodes are generally coated with some type of dielectric material to maintain a low surface area of exposed metal, yielding a high impedance. This high impedance enables the detection of the weak currents (order tens of nA) associated with single neuron action potentials with voltages on the scale of tens to hundreds of microvolts Carter and Shieh (2015). The shafts of the Utah electrode array are insulated with parylene-C (Blackrock Neurotech, Salt Lake City, UT). A small area at the tip of each electrode remains uncoated, through which the ionic current from nearby neurons can be measured. During electrolytic lesioning, applied voltages can be large enough to etch off the dielectric coating from the shaft of the two electrodes used to lesion. As this coating is etched, the impedance of those electrodes falls. Using a constant voltage source to lesion could deliver an inappropriately large amount of current into the brain once the dielectric coating was etched off and impedance reduced, resulting in uncontrolled tissue damage. A constant current source is robust to this changing impedance, maintaining the desired electrical current.

Single unit electrodes are commonly manufactured to have an impedance on the order of 100 kΩ to 1 MΩ Maynard et al. (1997). In past electrolytic lesioning studies, the currents used to create lesions with corresponding behavioral deficits were on the order of 100 µA Glees and Cole (1950); Wurtz and Goldberg (1972). Therefore, in order to supply this current in the face of the high impedance dielectric coating associated with single unit electrodes, high tens to low hundreds of volts are needed, depending on the lesioning parameters chosen. Even though commercial devices exist for electrolytic lesioning (e.g., Ugo Basile, Gemonio, IT), the power supplied may not be sufficient to recreate some current amplitudes used to lesion in literature Wurtz and Goldberg (1972) and generally only support DC waveforms.

We designed and built a custom current source circuit (Fig. 1) that supplies this voltage while maintaining precision in the delivered current (precision error not exceeding 10 µA). The circuit features a commercial power operational amplifier, PA97 (Apex Microtechnology, Tucson, AZ), in a negative feedback loop DePaola (2020), implemented in accordance with the corresponding evaluation kit, EK28 (Apex Microtechnology, Tucson, AZ). Power is provided through a 12 V external power supply, and a boost converter is used to create a VCC and -VCC of 450 V and −450 V, respectively. These are both low-cost, off-the-shelf, readily-available discrete circuits. The amplitude of the current supplied through between the cathode and anode to the brain (shown as the load, RL) is tuned through altering the value of a variable potentiometer (RS ). The voltage supplied to the op-amp can also be altered by altering the bias resistance with the potentiometer, RB. The total cost of the parts needed is approximately three hundred US dollars, making it an affordable addition to a neuroelectrophysiology recording setup.

Experimental Setup

The lesioning device is built as a contained box, to which the power supply, measurement devices, and microelectrode array are externally connected (Figure 1 - figure supplement 1). An external switch on the lesioning device controls the power to the boost converter. This is the switch that is manually turned on and off to control the duration of current delivery for lesioning. The lesioning device is connected to two electrodes from the intracortically implanted microelectrode array (RL). All lesioning was performed using Utah electrode arrays with the same specifications (Blackrock Neurotech, Salt Lake City, UT). The arrays are 4 mmx4 mm, with 96 channels. Electrode shafts are made of silicon (Si) with a metallic outer layer (platinum or platinum-iridium), and coated with parylene-C. Electrodes are 1 mm in length and have a 400 µm inter-electrode spacing. The array is connected to the lesioning device with the array’s external CerePort (Blackrock Neurotech, Salt Lake City, UT) and a CerePort breakout adaptor that connects to specific electrodes from the array. A Fluke 179 True-RMS Digital Multimeter (Fluke Corporation, Everett, WA) is used as an ammeter in series before RL to measure the current passing through RL (the desired output). A voltmeter spans RL to measure the voltage delivered. The variable potentiometer RS is tuned by a dial on the outside of the lesioning device, allowing quick and easy calibration of the amplitude of the current. All animal procedures and protocols were approved by the Stanford University Institutional Animal Care and Use Committee (#D16-00134).

Ex Vivo Ovine and Porcine Testing

Initial testing of lesion parameters was performed ex vivo in three sheep and eight pig brains (Supplementary File 1). These un-fixed brains were from animals that were euthanized earlier that day for unrelated research and/or teaching efforts, in line with the reduction principle of animal research National Research Council (US) Committee on Guidelines for the Use of Animals in Neuroscience and Behavioral Research (2003); National Research Council (US) Committee for the Update of the Guide for the Care and Use of Laboratory Animals (2011). As these brains were from unrelated research efforts, we were not given the sex or exact age of the animals. The brain was kept moist with saline throughout the procedure. A Utah array was implanted with a pneumatic inserter (Blackrock Neurotech, Salt Lake City, UT) into suitably large and flat gyri. The lesioning device was then connected to the microelectrode array and used to create a lesion. The location of the array implantation was marked using a histology pen or ink injection and detailed for the research record. The brains were then fixed, sliced, and prepared with hematoxylin and eosin (H&E) staining.

In Vivo Anesthetized Porcine Testing

We next sought to determine how lesion size might be altered in vivo, where blood flow, microhemorrhage, and the body’s acute inflammatory responses could affect the lesion extent and intensity over time. Testing was performed on five anesthetized pigs, to evaluate the ex vivo current amplitude and duration parameters in the context of inflammatory responses (Supplementary File 2).

Each pig was sedated and then intubated, ventilated, and placed on inhaled isolflurane. Once under anesthesia and the airway was confirmed, the head was secured in preparation for the procedure. After preparing and cleaning the surgical area, a midline skin incision was made and all skin, muscle, aponeurosis, and periosteum were retracted to expose the skull. Most of the dorsal surface of the skull was exposed along with some of the lateral margins to visualize the anatomy. A single, large craniectomy, exposing most of the superior surface of the dura, was made using a high speed bone drill (ANSPACH, DePuy Synthes, Raynham, MA). Once the craniectomy was at the desired size and location, the dura was washed with saline and hemisphere-wide dural flaps were made to expose the brain. Once exposed, the brain was covered with gauze and kept moist with saline throughout the procedure.

The implantation and lesion procedures were completed in the same manner as in the ex vivo testing. After lesioning, the array was either removed by hand and placed at a new cortical location for further testing or, if the last lesion of the procedure, was left in place for two to three hours before euthanizing the animal to even better mimic the post-lesion inflammatory response. Therefore, histology for all lesions included the physiological response to lesioning after at least two to three hours, and some lesions also retained the array in place as a foreign body during this post-lesion time frame. After the final lesioning was complete and sufficient time had elapsed for lesions to appear histologically, the pig was euthanized via IV injection of beauthanasia solution (100mg/kg, IV). The brain was then fixed, and slices were prepared with H&E staining.

In Vivo Lesions in Rhesus Macaque

Electrolytic lesioning is not painful as there are no direct pain receptors in the central nervous system Hall and Hall (2021). Based on this, and a lack of physiological signs of pain from the anaesthetized pig studies, a lesion was performed on a sedated rhesus macaque who was subsequently euthanized due to unrelated health complications (Monkey F; 16 year-old adult, male rhesus macaque) in order to further verify safety before use in awake-behaving rhesus. This lesion was created by applying 150 µA of direct current to two adjacent electrodes in the microelectrode array for 45 seconds. Again, no physiological signs of stress or complications were observed, increasing confidence for lesions in awake animals. Subsequently, lesions were safely performed in two other rhesus (Monkeys H and U; 14 and 11 years-old, respectively, adult, male rhesus macaques) while awake and seated in a primate chair, without the animals exhibiting any behavioral signs of stress or pain. Across these two animals, fourteen lesions were performed using thirteen unique electrode pairs, demonstrating that the same microelectrode array can be used for several lesions and even the same electrode pair can be used as the anode and cathode for multiple lesions. In these lesions, 150 µA of direct current was applied to two adjacent electrodes in the microelectrode array for 30 or 45 seconds (30s for Monkey U, 45s for Monkey H), except in lesion H200120 where current was applied for 50 seconds. The Utah array chronically implanted in the monkey’s cortex is connected to the lesioning device via its CerePort (the same one used for recording the neuronal signals from the microelectrode array). The lesioning device is then turned on for a fixed duration, delivering the desired current. The animal is continuously monitored during the procedure for any signs of discomfort. After lesioning is complete and the recording cables are re-connected to the CerePort, the monkey is immediately ready to resume electrophysiological recordings and participate in a wide variety of head-fixed or freely moving behavioral paradigms. Although lesioning itself is painless, the technique is intended to cause a temporary functional impairment. In light of this, monitoring of animal behavior post-procedure is conducted by research staff in coordination with veterinary staff to ensure health, safety, and psychological well-being.

Estimation of Lesion Volume

In order to explore the relationship between the amplitude and duration of the applied current and lesion volume, we needed to have estimates of lesion volume. Gathering multiple histological slices per lesion would have been cost prohibitive, so we attempted to take a histology slice from the estimated center of the lesion. If the histology slice successfully contained the lesion, we assumed the slice bisected the lesion volume. Occasionally, however, the histology slice did not contain the lesion. For these lesions, if we did have a macro camera photo and the lesion was large enough to create a visible cavitation, we used this photo to estimate the diameter of the lesion, and as no information was available about the height of the lesion cavitation, we estimated the height to be equivalent to the radius. We used the radius and height (gathered either from histology or a photo) to estimate the volume of each lesion, using volume formulas for the approximate geometries observed visually in the histology. Most volumes were estimated assuming a conical lesion volume, which was chosen based on Figure 2 and Figure 3. Only one in vivo lesion, where 150 µA direct current was applied for 30 s, was calculated as a spherical volume as indicated by the visually identified damage in the histology slice. When we plotted just lesions created with a direct current amplitude of 180 µA, we fit the data with an exponential curve (curve_fit, SciPy optimize).

Neuronal Recordings and Processing

Neuronal data was collected at 30 kHz from primary motor cortex of Monkeys H and U using a Utah array (described above), captured using a Cerebus system (Blackrock Neurotech, Salt Lake City, UT), then high-pass filtered offline with a fourth order, zero-phase Butterworth filter using a 250 Hz cutoff frequency. Action potentials were identified by thresholding the filtered membrane potential at −4× the root mean square voltage measured over the first minute of the recording session. Data snippets for comparing electrophysiology pre- and post-lesion were generated by taking the preceding 16 sampled data points (0.53 ms), and proceeding 32 sampled points (1.07 ms) around the peak depolarization identified after each threshold crossing. Waveforms whose peak-to-trough amplitude exceeded 300 µV are atypical of extracellular recordings from cortical neurons using Utah arrays and were excluded Chestek et al. (2011). Waveform width was calculated as the difference between the two prominent peaks in the temporal derivative of the extracellular potential. Unphysiological waveforms whose widths were longer than 1 ms were also excluded Bean (2007). The remaining identified action potential waveforms were then averaged for each electrode, and event rates were determined as the number of spikes on a given electrode divided by the duration of the recording session (excluding brief periods of inactivity before the start of the session).

As a putative indication of neuronal loss, neuron turnover was measured between daily recording sessions (each pair separated by at most three days, a time frame that minimizes natural turnover in recorded waveforms Gallego et al. (2020)). For a given electrode, 1000 identified action potential samples (each 1.6 ms, 48 points sampled at 30 kHz from the first day’s recording session were peak-aligned to create a 1000×48 matrix, where each row contains the 48 time points of a given waveform sampled at 30 kHz. Each of these points in time were considered as a sample feature and principal component analysis was performed. Next, the second day’s waveforms were projected onto the top two principal components of the first day (see Figure 7B). The distance of each projected waveform from the origin was calculated for each day; a median radius value was computed and represented as a circle around the origin. Changes in the radius of these circles, Δr, captures changes in the composition of multi-unit activity recorded at each electrode. By fitting Gaussian mixture models to the distributions of Δr, we were able to find evidence of discrete clusters of radial changes. One and two standard deviations from the median cluster value were used to define five categories: matching waveforms, small increase, small decrease, large increase, and large decrease. A given electrode’s waveforms were said to match if Δr ≈ 0 and was considered altered otherwise. Electrodes with matching waveforms are presented as a percentage of the array’s total 96 electrodes for a given pair of days (% match).

These pairwise comparisons were then split up into three groups: pre-lesion versus pre-lesion days (pre-pre), pre-lesion versus post-lesion days (pre-post), and post-lesion version post-lesion days (post-post). Although a set of eleven consecutive days was analyzed for each lesion (four pre- and seven post-), only recording session pairs that were separated by three-days or fewer were analyzed in order to minimize the potential confound of increased neuron turnover resulting from longer separation between sessions Gallego et al. (2020). Comparisons were then made between the pre-lesion period, an acute injury period (up to three days following a lesion), and a late injury period (days four to seven following a lesion). The Median test was used to test the null hypothesis that lesioning had no effect on recorded turnover with 95% confidence interval, whose significance threshold was Bonferroni corrected for the three statistical comparisons made between groups (α∕3, significant if p < 0.017).

Ex Vivo Testing with a Linear Multielectrode Probe

Testing of the ability to lesion through a different type of multielectrode probe (a linear multielectrode probe, rather than the microelectrode arrays used throughout the rest of the study) was done in an ex vivo rabbit brain. The un-fixed brain was from an animal that was euthanized earlier that day for unrelated research efforts. As this brain was from an unrelated research effort, we were not given the sex or age of the animal. The brain was kept moist with saline throughout the procedure. A U-Probe (Plexon, Dallas, TX), with 24 contacts, 100 µm spacing between them, and an impedance of 1-1.5 MΩ, was lowered one cm into cortex using a stereotaxic manipulator. Using a modified connector that attached to the female connector at the top of the probe, we connected our electrolytic lesioning device to two of the contacts on the probe, to act as the anode and cathode through which current was passed. The two contacts were contiguous (100 µm apart), and the lesions were created at depths approximately 400 µm below the surface of cortex. The location of the probe insertion was marked using ink injection, and is indicated in a sketch in Figure 8A.

Supporting information